---
title: "eb_ica"
author: "junmingguan"
date: "2026-03-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

We examine an empirical Bayes version of ICA, where the prior $p_\pi(s_i)=\text{Rad}(s_i;\ \pi)$ is estimated from the data. The steps to find one source are as follows:

-   M-step: $\mathbf{w} = Y \mathbf{s}^+ / \| Y \mathbf{s}^+ \|$

-   E-step:

    $$
    \begin{align}
    (\hat{q}(\mathbf{s}), \hat{\pi}) &=\arg \max_{q, \pi} \mathbb{E}_q\left[\log p(\mathbf{s}, Y \mid \mathbf{w})\right] \\
    \\
    &= \arg \max_{q, \pi} \mathbb{E}_q\left[ \log p(Y \mid \mathbf{s}, \mathbf{w}) + \sum \log p_\pi(s_i )\right] \\
    &= \arg \max_{q, \pi}  \left\{-\frac{1}{2\sigma^2} (-2 \text{tr}(\mathbb{E}_q[\mathbf{s}]^\top Y^\top \mathbf{w}) + \mathbb{E}_q\|\mathbf{s}\|^2) -\sum\text{KL}\Big(q(s_i)\ \|\ p_\pi(s_i)\Big)\right\} 
    \end{align}
    $$

    -   solved by $\text{EBNM}(Y^\top\mathbf{w}, \sigma)$
    -   $\mathbf{s}^+=\mathbb{E}_\hat{q}[\mathbf{s}]$.

For a rank-K fit, we proceed as follows:

-   M-step:

    -   $W_+ = YS^+$

    -   Orthogonalization: $W = U_+ V_+^\top$, where $W^+=U_+ DV_+^\top$, which is equivalent to $W= (W_+W_+^\top)^{-1/2} W_+$ proposed in [@Karhunenetal].

-   E-step: for $k=1,\dots, K$,

    -   Obtain $\hat{q}_k, \hat{\pi}_k$ by $\text{EBNM}(Y^\top\mathbf{w}_k, \sigma)$

    -   $\mathbf{s}_k^+=\mathbb{E}_\hat{q}[\mathbf{s}_k]$.

```{r, warning=FALSE, message=FALSE}
library(flashier)
library(Matrix)
library(fastICA)
library(fastTopics)
library(ebnm)
source('code/ebcd_functions.R')
source('code/ica_functions.R')
source('code/ebnm_rademacher.R')
```

```{r}
simulate_unbalanaced_rad_groups <- function(
    K, n=100, p=1000, noise=1, percent_one=0.2) {
  set.seed(1)
  L = matrix(-1,nrow=n,ncol=K)
  for(i in 1:K){L[sample(1:n,floor(percent_one*n)),i]=1}
  FF = matrix(rnorm(p*K), nrow = p, ncol=K)
  X = L %*% t(FF) + rnorm(n*p, 0, noise)
  return(list(X=X, L=L, FF=FF))
}
```

```{r}
sim_data <- simulate_unbalanaced_rad_groups(
  K=3, n=100, p=1000, percent_one = 0.2, noise = 1
)
X   <- sim_data$X
L   <- sim_data$L
FF  <- sim_data$FF
X1  <- preprocess(X)
```

## Unbalanced 3 groups

### rank 1 fit

```{r}
cor_max_ebica   <- c() # maximum correlation between the estmate and a true source
cor_max_fastica <- c() # maximum correlation between the estmate and a true source

est_S_ebica   <- matrix(0, nrow = nrow(X), ncol = 100) # all 100 solutions
est_S_fastica <- matrix(0, nrow = nrow(X), ncol = 100) # all 100 solutions

for (i in 1:100) {
  set.seed(i)
  s <- matrix(rnorm(nrow(X)), ncol = 1)
  w <- X1 %*% s
  ebica_res <- ebica(t(X), K=1, 
                       S_init = s)
  fastica_w <- fastica_r1update(w, X1)
  fastica_s <- t(X1) %*% fastica_w
  cor_max_ebica <- c(cor_max_ebica, max(abs(cor(L, ebica_res$S_plus))))
  est_S_ebica[,i] <- ebica_res$S_plus
  cor_max_fastica <- c(cor_max_fastica, max(abs(cor(L, fastica_s))))
  est_S_fastica[,i] <- fastica_s
  
}
table(abs(cor_max_ebica) > 0.99)
table(abs(cor_max_fastica) > 0.99)
image(t(est_S_ebica[, order((cor_max_ebica), decreasing = T)]))
image(t(est_S_fastica[, order((cor_max_fastica), decreasing = T)]))
```

### rank 3 fit

```{r}
K <- 3
ebica_res <- ebica(t(X), K=K, 
             S_init = NULL)
cor(L, ebica_res$S_plus)



fastica_res <- fastICA(X, n.comp = K, alg.typ = "parallel")
cor(L, fastica_res$S)
```

### rank 4 fit

```{r}
K <- 4
ebica_res <- ebica(t(X), K=K, 
             S_init = NULL)
cor(L, ebica_res$S_plus)



fastica_res <- fastICA(X, n.comp = K, alg.typ = "parallel")
cor(L, fastica_res$S)
```

## Unbalanced 9 groups

```{r}
sim_data <- simulate_unbalanaced_rad_groups(
  K=9, n=100, p=1000, percent_one = 0.2, noise = 1
)
X   <- sim_data$X
L   <- sim_data$L
FF  <- sim_data$FF
X1  <- preprocess(X)
```

```{r}
set.seed(1)
K <- 9
ebica_res <- ebica(t(X), K=K, 
             S_init = NULL)
cor(L, ebica_res$S_plus)
apply(abs(cor(L, ebica_res$S_plus)), 1, max)


fastica_res <- fastICA(X, n.comp = K, alg.typ = "parallel")
cor(L, fastica_res$S)
apply(abs(cor(L, fastica_res$S)), 1, max)
```
