---
title: "Rank sensitivity of EB projection methods and FastICA"
author: "Junming Guan"
date: "2026-07-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal


```{r include-estimator-code, include=FALSE, purl=FALSE}
knitr::read_chunk(file.path("code", "ica_functions.R"))
knitr::read_chunk(file.path("code", "ebproj_functions.R"))
```

## Setup

```{r}
source(file.path("code", "ebnm_rademacher.R"))
source(file.path("code", "ebnm_shifted_rademacher.R"))
source(file.path("code", "ica_functions.R"))
source(file.path("code", "ebproj_functions.R"))

n_starts <- 20
kappa <- 1
sigma <- 0.1
k_pca <- 20
power_tilt_gap <- 0.01
```



## Experiment runner

The reference simulations use

$$
X = A S + E,\qquad
A_{jk} \sim N(0,1),\quad E_{ij}\sim N(0,0.1^2).
$$

Here $X$ is stored as features by samples in the reference document and
transposed below so our estimators receive samples by features. The first
test uses $n=200$; Tests 2--4 use $n=100$; all tests use $p=1000$. The
reference retains 20 PCs, so our rank grids start at the true signal rank and
end at 20. Increasing $K$ therefore only adds noise PCs.

```{r}
recovery_correlation <- function(v, S) {
  if (!all(is.finite(v)) || sd(v) == 0) {
    return(0)
  }
  max(abs(cor(v, S)))
}

run_rank_test <- function(X, S, k_grid, test_name, ebnm_fn) {
  X <- scale(X, center = TRUE, scale = FALSE)
  n_samples <- nrow(X)
  set.seed(1)
  v_inits <- replicate(
    n_starts,
    sample(rep(c(-1, 1), length.out = n_samples)),
    simplify = FALSE
  )

  sv <- svd(X, nu = max(k_grid), nv = 0)
  results <- data.frame()

  for (K in k_grid) {
    U <- sv$u[, seq_len(K), drop = FALSE]
    P <- tcrossprod(U)
    d <- sv$d[seq_len(K)]
    W <- tcrossprod(sweep(U, 2, d / d[1], "*"))
    W_power <- (1 - power_tilt_gap) * W
    Z_whitened <- sqrt(n_samples) * t(U)
    Z_nonwhitened <- sweep(t(U), 1, d, "*")

    for (start in seq_len(n_starts)) {
      w_init <- as.vector(crossprod(U, v_inits[[start]]))

      fit_ebproj <- eb_proj_rank1_update(
        P = P,
        v_init = v_inits[[start]],
        ebnm_fn = ebnm_fn,
        k = K,
        b = 1,
        maxiter = 100
      )

      fit_spiked <- eb_spiked_proj_rank1_update(
        P = P,
        v_init = v_inits[[start]],
        ebnm_fn = ebnm_fn,
        k = K,
        kappa = kappa,
        maxiter = 100
      )

      fit_weighted <- eb_weighted_proj_rank1_update(
        W = W,
        eta = n_samples / 2,
        m_init = v_inits[[start]],
        M_init = sum(v_inits[[start]]^2),
        ebnm_fn = ebnm_fn,
        maxiter = 100
      )

      fit_power_tilt <- eb_weighted_proj_pow_tilt_rank1_update(
        W = W_power,
        gamma = (n_samples - K - 2) / 2,
        m_init = v_inits[[start]],
        M_init = sum(v_inits[[start]]^2),
        ebnm_fn = ebnm_fn,
        maxiter = 100
      )

      w_fastica <- fastica_rank_one_update(
        X1 = Z_whitened,
        init_w = w_init,
        n_iter = 100
      )

      w_fastica_nw <- fastica_rank_one_update_nonwhitened(
        Z = Z_nonwhitened,
        d2 = d^2,
        init_w = w_init,
        n_iter = 100
      )

      cor_ebproj <- recovery_correlation(fit_ebproj$vbar, S)
      cor_spiked <- recovery_correlation(fit_spiked$vbar, S)
      cor_weighted <- recovery_correlation(fit_weighted$vbar, S)
      cor_power_tilt <- recovery_correlation(fit_power_tilt$vbar, S)
      cor_fastica <- recovery_correlation(
        as.vector(t(Z_whitened) %*% w_fastica),
        S
      )
      cor_fastica_nw <- recovery_correlation(
        as.vector(t(Z_nonwhitened) %*% w_fastica_nw),
        S
      )

      results <- rbind(
        results,
        data.frame(
          test = test_name,
          rank = K,
          start = start,
          method = c(
            "EB-proj",
            "EB-spiked-proj",
            "EB-weighted-proj",
            "EB-weighted-proj-pow-tilt",
            "FastICA (whitened)",
            "FastICA (non-whitened)"
          ),
          correlation = c(
            cor_ebproj,
            cor_spiked,
            cor_weighted,
            cor_power_tilt,
            cor_fastica,
            cor_fastica_nw
          )
        )
      )
    }
  }

  results$success <- results$correlation > 0.9
  list(
    results = results,
    singular_values = sv$d,
    k_grid = k_grid,
    true_rank = ncol(S)
  )
}
```

## Test 1: one symmetric binary source

```{r}
set.seed(10)
n1 <- 200
p1 <- 1000
k_true1 <- 1
A1 <- matrix(rnorm(p1 * k_true1), nrow = p1)
S_rad <- matrix(sample(c(-1, 1), n1, replace = TRUE), nrow = 1)
X_rad <- A1 %*% S_rad +
  matrix(rnorm(p1 * n1, sd = sigma), nrow = p1)

test1 <- run_rank_test(
  X = t(X_rad),
  S = t(S_rad),
  k_grid = c(1, 2, 5, 10, 15, k_pca),
  test_name = "Symmetric binary source",
  ebnm_fn = ebnm_rademacher
)
```

## Test 2: four non-overlapping groups

Each source selects one group of 25 observations. After centering, each source
has a shifted binary distribution, so we use the shifted Rademacher EBNM
family for all three EB methods.

```{r}
set.seed(1)
n2 <- 100
p2 <- 1000
k_true2 <- 4
A2 <- matrix(rnorm(p2 * k_true2), nrow = p2)
S_groups <- matrix(0, nrow = k_true2, ncol = n2)
S_groups[1, 1:25] <- 1
S_groups[2, 26:50] <- 1
S_groups[3, 51:75] <- 1
S_groups[4, 76:100] <- 1
X_groups <- A2 %*% S_groups +
  matrix(rnorm(p2 * n2, sd = sigma), nrow = p2)

test2 <- run_rank_test(
  X = t(X_groups),
  S = t(S_groups),
  k_grid = c(4, 5, 10, 15, k_pca),
  test_name = "Four non-overlapping groups",
  ebnm_fn = ebnm_generalized_shifted_rademacher
)
```

## Test 3: four sparse binary sources

Each source is active in 20 randomly chosen observations, independently across
sources.

```{r}
set.seed(2)
n3 <- 100
p3 <- 1000
k_true3 <- 4
prob_active <- 0.2
A3 <- matrix(rnorm(p3 * k_true3), nrow = p3)
S_sparse <- matrix(0, nrow = k_true3, ncol = n3)
for (j in seq_len(k_true3)) {
  S_sparse[j, sample(n3, round(prob_active * n3))] <- 1
}
X_sparse <- A3 %*% S_sparse +
  matrix(rnorm(p3 * n3, sd = sigma), nrow = p3)

test3 <- run_rank_test(
  X = t(X_sparse),
  S = t(S_sparse),
  k_grid = c(4, 5, 10, 15, k_pca),
  test_name = "Four sparse sources",
  ebnm_fn = ebnm_generalized_shifted_rademacher
)
```

## Test 4: nine overlapping groups

This is the larger overlapping-group setting from the reference experiment.
Each of the nine sources is active in 20 observations.

```{r}
set.seed(1)
n4 <- 100
p4 <- 1000
k_true4 <- 9
L <- matrix(0, nrow = n4, ncol = k_true4)
for (j in seq_len(k_true4)) {
  L[sample(n4, 20), j] <- 1
}
FF <- matrix(rnorm(p4 * k_true4), nrow = p4, ncol = k_true4)
X9 <- t(
  L %*% t(FF) +
    matrix(rnorm(n4 * p4, sd = sigma), nrow = n4)
)

test4 <- run_rank_test(
  X = t(X9),
  S = L,
  k_grid = c(9, 10, 15, k_pca),
  test_name = "Nine overlapping groups",
  ebnm_fn = ebnm_generalized_shifted_rademacher
)
```

## Check the signal and noise PCs

```{r, fig.width=8, fig.height=7}
tests <- list(test1, test2, test3, test4)
test_names <- vapply(tests, function(x) x$results$test[1], character(1))

old_par <- par(mfrow = c(2, 2))
for (i in seq_along(tests)) {
  x <- tests[[i]]
  plot(
    x$singular_values[seq_len(max(x$k_grid))],
    type = "b",
    pch = 19,
    xlab = "PC index",
    ylab = "Singular value",
    main = test_names[i]
  )
  abline(v = x$true_rank + 0.5, lty = 2, col = "grey50")
}
par(old_par)
```

## Summarize rank sensitivity

```{r, results='asis'}
results <- do.call(rbind, lapply(tests, `[[`, "results"))

mean_cor <- aggregate(
  correlation ~ test + rank + method,
  data = results,
  FUN = mean
)
names(mean_cor)[4] <- "mean_correlation"

success_rate <- aggregate(
  success ~ test + rank + method,
  data = results,
  FUN = mean
)

summary_results <- merge(mean_cor, success_rate)
summary_results <- summary_results[
  order(summary_results$test, summary_results$rank, summary_results$method),
]

for (test_name in test_names) {
  test_table <- summary_results[
    summary_results$test == test_name,
    c("rank", "method", "mean_correlation", "success")
  ]

  cat("\n\n### ", test_name, "\n\n", sep = "")
  cat(
    knitr::kable(
      test_table,
      format = "pipe",
      digits = 3,
      row.names = FALSE
    ),
    sep = "\n"
  )
  cat("\n\n")
}
```
