---
title: "Real data experiments"
author: "Junming Guan"
date: "2026-07-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

```{r}
library(Matrix)
library(fastTopics)
library(DuoClustering2018)
library(irlba)

source(file.path("code", "ebnm_rademacher.R"))
source(file.path("code", "ebnm_shifted_rademacher.R"))
source(file.path("code", "ebproj_functions.R"))
source(file.path("code", "ica_functions.R"))

set.seed(3)
```

# PBMC FACS

## Prepare the data

This follows the preprocessing in the
[flashier single-cell vignette](https://willwerscheid.github.io/flashier/articles/flashier_single_cell.html).

```{r}
data("pbmc_facs")

counts <- pbmc_facs$counts
colnames(counts) <- make.names(pbmc_facs$genes$symbol, unique = TRUE)

a <- 1
size_factors <- rowSums(counts)
size_factors <- size_factors / mean(size_factors)
shifted_log_counts <- log1p(counts / (a * size_factors))

dim(shifted_log_counts)
```

## Construct the projection

Center the shifted-log data, use the leading eight left singular vectors, and
form the cell-space projection matrix \(P=UU^\top\).

```{r}
K <- 8
gene_means <- colMeans(shifted_log_counts)
svd_fit <- irlba(
  shifted_log_counts,
  nv = K,
  nu = K,
  center = gene_means
)
U <- svd_fit$u
P <- tcrossprod(U)
```

## Run EB-proj

Initialize from the sign of the first centered singular vector and run one
rank-one EB-proj update with the symmetric Rademacher prior.

```{r}
v_init <- sign(U[, 1])
v_init[v_init == 0] <- 1

fit_ebproj <- eb_proj_rank1_update(
  P = P,
  v_init = v_init,
  ebnm_fn = ebnm_generalized_shifted_rademacher,
  k = K,
  b = 1,
  maxiter = 100
)

c(rho = fit_ebproj$rho, niter = fit_ebproj$niter)
fit_ebproj$fitted_g
```

## Run EB-spiked-proj

Use the same initialization and Rademacher prior, with \(\kappa=1\).

```{r}
fit_ebspikedproj <- eb_spiked_proj_rank1_update(
  P = P,
  v_init = v_init,
  ebnm_fn = ebnm_generalized_shifted_rademacher,
  k = K,
  kappa = 1,
  maxiter = 100
)

c(
  rho = fit_ebspikedproj$rho,
  niter = fit_ebspikedproj$niter,
  eta = fit_ebspikedproj$eta,
  kappa = fit_ebspikedproj$kappa
)
```

## Run FastICA

The whitened cell matrix is \(X_1=\sqrt{n}U^\top\). Use the same initial
cell direction as EB-proj.

```{r}
n <- nrow(U)
X1 <- sqrt(n) * t(U)

w_init <- as.vector(X1 %*% v_init)
w_init <- w_init / sqrt(sum(w_init^2))

w_fastica <- fastica_rank_one_update(
  X1 = X1,
  init_w = w_init,
  n_iter = 100
)
score_fastica <- as.vector(t(X1) %*% w_fastica)
```

## Run FastICA with an intercept

Add an all-ones row to the whitened data and estimate its coefficient jointly
with the FastICA direction.

```{r}
X1_intercept <- rbind(intercept = rep(1, ncol(X1)), X1)
w_init_intercept <- c(0, w_init)

w_fastica_intercept <- fastica_rank_one_update(
  X1 = X1_intercept,
  init_w = w_init_intercept,
  n_iter = 100
)
score_fastica_intercept <- as.vector(
  t(X1_intercept) %*% w_fastica_intercept
)

c(intercept_weight = w_fastica_intercept[1])
```

## Compare the fitted contrasts

Signs and scales are arbitrary. Align the other scores to EB-proj and
standardize all four before comparing them.

```{r}
orient_and_scale <- function(x, reference) {
  if (cor(x, reference) < 0) {
    x <- -x
  }
  as.vector(scale(x))
}

scores <- list(
  "EB-proj" = as.vector(scale(fit_ebproj$vbar)),
  "EB-spiked-proj" = orient_and_scale(
    fit_ebspikedproj$vbar,
    fit_ebproj$vbar
  ),
  "FastICA" = orient_and_scale(score_fastica, fit_ebproj$vbar),
  "FastICA + intercept" = orient_and_scale(
    score_fastica_intercept,
    fit_ebproj$vbar
  )
)

subpop <- pbmc_facs$samples$subpop
round(sapply(scores, function(x) tapply(x, subpop, mean)), 2)
round(cor(do.call(cbind, scores)), 2)

old_par <- par(mfrow = c(2, 2), mar = c(8, 4, 3, 1))
for (method in names(scores)) {
  boxplot(
    scores[[method]] ~ subpop,
    main = method,
    xlab = "",
    ylab = "Standardized score",
    las = 2
  )
}
par(old_par)
```

At \(\kappa=1\), EB-spiked-proj has \(\eta=0.9997\) and recovers
essentially the same direction as EB-proj (score correlation 1.00). Both place
CD14+ and NK cells on one side, B and CD34+ cells on the other, and T cells
near the center. Their score correlations are 0.56 with ordinary FastICA and
0.70 with FastICA plus an intercept.

# Zhengmix4eq

## Prepare the data

`Zhengmix4eq` contains four pre-sorted immune-cell populations mixed in
approximately equal proportions. Use the version retaining the top 10% of
highly variable genes; all cells and their pre-sorting labels are retained.

```{r}
sce <- sce_filteredHVG10_Zhengmix4eq()
counts <- t(SummarizedExperiment::assay(sce, "counts"))
cell_type <- factor(sce$phenoid)

dim(counts)
table(cell_type)

size_factors <- rowSums(counts)
size_factors <- size_factors / mean(size_factors)
shifted_log_counts <- log1p(counts / size_factors)
```

## Construct the projection

Center genes, retain the first eight left singular vectors, and construct the
cell-space projection \(P=UU^\top\).

```{r}
K <- 8
svd_fit <- irlba(
  shifted_log_counts,
  nv = K,
  nu = K,
  center = colMeans(shifted_log_counts)
)
U <- svd_fit$u
P <- tcrossprod(U)
```

## Run EB-proj

Initialize from the sign of the first centered singular vector and use the
generalized shifted Rademacher prior.

```{r}
v_init <- sign(U[, 1])
v_init[v_init == 0] <- 1

fit_ebproj <- eb_proj_rank1_update(
  P = P,
  v_init = v_init,
  ebnm_fn = ebnm_generalized_shifted_rademacher,
  k = K,
  b = 1,
  maxiter = 100
)

c(rho = fit_ebproj$rho, niter = fit_ebproj$niter)
fit_ebproj$fitted_g
```

## Run EB-spiked-proj

Use the same initialization and prior, with \(\kappa=1\).

```{r}
fit_ebspikedproj <- eb_spiked_proj_rank1_update(
  P = P,
  v_init = v_init,
  ebnm_fn = ebnm_generalized_shifted_rademacher,
  k = K,
  kappa = 1,
  maxiter = 100
)

c(
  rho = fit_ebspikedproj$rho,
  niter = fit_ebspikedproj$niter,
  eta = fit_ebspikedproj$eta,
  kappa = fit_ebspikedproj$kappa
)
fit_ebspikedproj$fitted_g
```

## Run FastICA

Use the whitened cell matrix \(X_1=\sqrt{n}U^\top\) and the same initial
direction.

```{r}
n <- nrow(U)
X1 <- sqrt(n) * t(U)

w_init <- as.vector(X1 %*% v_init)
w_init <- w_init / sqrt(sum(w_init^2))

w_fastica <- fastica_rank_one_update(
  X1 = X1,
  init_w = w_init,
  n_iter = 100
)
score_fastica <- as.vector(t(X1) %*% w_fastica)
```

## Run FastICA with an intercept

```{r}
X1_intercept <- rbind(intercept = rep(1, ncol(X1)), X1)
w_init_intercept <- c(0, w_init)

w_fastica_intercept <- fastica_rank_one_update(
  X1 = X1_intercept,
  init_w = w_init_intercept,
  n_iter = 100
)
score_fastica_intercept <- as.vector(
  t(X1_intercept) %*% w_fastica_intercept
)

intercept_diagnostics <- c(
  intercept_weight = w_fastica_intercept[1],
  direction_norm = sqrt(sum(w_fastica_intercept[-1]^2)),
  raw_score_sd = sd(score_fastica_intercept)
)
intercept_diagnostics

intercept_degenerate <- intercept_diagnostics["direction_norm"] < 1e-8
```

## Compare with the pre-sorting labels

Signs and scales are arbitrary. Align the other scores to EB-proj and
standardize all nondegenerate methods. `label_R2` is the fraction of each
score's variance explained by the four known cell labels.

```{r}
scores <- list(
  "EB-proj" = as.vector(scale(fit_ebproj$vbar)),
  "EB-spiked-proj" = orient_and_scale(
    fit_ebspikedproj$vbar,
    fit_ebproj$vbar
  ),
  "FastICA" = orient_and_scale(score_fastica, fit_ebproj$vbar)
)
if (!intercept_degenerate) {
  scores[["FastICA + intercept"]] <- orient_and_scale(
    score_fastica_intercept,
    fit_ebproj$vbar
  )
}

round(sapply(scores, function(x) tapply(x, cell_type, mean)), 2)
round(cor(do.call(cbind, scores)), 2)

label_R2 <- sapply(
  scores,
  function(x) summary(lm(x ~ cell_type))$r.squared
)
round(label_R2, 3)

old_par <- par(mfrow = c(2, 2), mar = c(8, 4, 3, 1))
for (method in names(scores)) {
  boxplot(
    scores[[method]] ~ cell_type,
    main = method,
    xlab = "",
    ylab = "Standardized score",
    las = 2
  )
}
par(old_par)
```

Because four labels span a three-dimensional contrast space, a single
rank-one component can recover only one split among the populations. Thus,
`label_R2` measures association with the known labels, not complete recovery
of the four-class structure. The intercept FastICA fit is excluded from these
comparisons if it converges to an intercept-only, constant-score solution.

With the generalized shifted Rademacher prior, both EB methods isolate CD14
monocytes from the other three populations and produce identical scores. Their
label \(R^2\) is 0.978, compared with 0.845 for FastICA, and the EB--FastICA
score correlation is 0.67. FastICA with an intercept converges to an
intercept-only solution and therefore has no usable cell score.

# Zhengmix4uneq

## Prepare the data

`Zhengmix4uneq` contains the same four pre-sorted immune-cell populations in
deliberately unequal proportions. Use the version retaining the top 10% of
highly variable genes; all retained cells and their pre-sorting labels are
included.

```{r}
sce <- sce_filteredHVG10_Zhengmix4uneq()
counts <- t(SummarizedExperiment::assay(sce, "counts"))
cell_type <- factor(sce$phenoid)

dim(counts)
table(cell_type)

size_factors <- rowSums(counts)
size_factors <- size_factors / mean(size_factors)
shifted_log_counts <- log1p(counts / size_factors)
```

## Construct the projection

Center genes, retain the first eight left singular vectors, and construct the
cell-space projection \(P=UU^\top\).

```{r}
K <- 8
svd_fit <- irlba(
  shifted_log_counts,
  nv = K,
  nu = K,
  center = colMeans(shifted_log_counts)
)
U <- svd_fit$u
P <- tcrossprod(U)
```

## Run EB-proj

Initialize from the sign of the first centered singular vector and use the
generalized shifted Rademacher prior.

```{r}
v_init <- sign(U[, 1])
v_init[v_init == 0] <- 1

fit_ebproj <- eb_proj_rank1_update(
  P = P,
  v_init = v_init,
  ebnm_fn = ebnm_generalized_shifted_rademacher,
  k = K,
  b = 1,
  maxiter = 100
)

c(rho = fit_ebproj$rho, niter = fit_ebproj$niter)
fit_ebproj$fitted_g
```

## Run EB-spiked-proj

Use the same initialization and prior, with \(\kappa=1\).

```{r}
fit_ebspikedproj <- eb_spiked_proj_rank1_update(
  P = P,
  v_init = v_init,
  ebnm_fn = ebnm_generalized_shifted_rademacher,
  k = K,
  kappa = 1,
  maxiter = 100
)

c(
  rho = fit_ebspikedproj$rho,
  niter = fit_ebspikedproj$niter,
  eta = fit_ebspikedproj$eta,
  kappa = fit_ebspikedproj$kappa
)
fit_ebspikedproj$fitted_g
```

## Run FastICA

Use the whitened cell matrix \(X_1=\sqrt{n}U^\top\) and the same initial
direction.

```{r}
n <- nrow(U)
X1 <- sqrt(n) * t(U)

w_init <- as.vector(X1 %*% v_init)
w_init <- w_init / sqrt(sum(w_init^2))

w_fastica <- fastica_rank_one_update(
  X1 = X1,
  init_w = w_init,
  n_iter = 100
)
score_fastica <- as.vector(t(X1) %*% w_fastica)
```

## Run FastICA with an intercept

```{r}
X1_intercept <- rbind(intercept = rep(1, ncol(X1)), X1)
w_init_intercept <- c(0, w_init)

w_fastica_intercept <- fastica_rank_one_update(
  X1 = X1_intercept,
  init_w = w_init_intercept,
  n_iter = 100
)
score_fastica_intercept <- as.vector(
  t(X1_intercept) %*% w_fastica_intercept
)

intercept_diagnostics <- c(
  intercept_weight = w_fastica_intercept[1],
  direction_norm = sqrt(sum(w_fastica_intercept[-1]^2)),
  raw_score_sd = sd(score_fastica_intercept)
)
intercept_diagnostics

intercept_degenerate <- intercept_diagnostics["direction_norm"] < 1e-8
```

## Compare with the pre-sorting labels

Signs and scales are arbitrary. Align the other scores to EB-proj and
standardize all nondegenerate methods. `label_R2` is the fraction of each
score's variance explained by the four known cell labels.

```{r}
scores <- list(
  "EB-proj" = as.vector(scale(fit_ebproj$vbar)),
  "EB-spiked-proj" = orient_and_scale(
    fit_ebspikedproj$vbar,
    fit_ebproj$vbar
  ),
  "FastICA" = orient_and_scale(score_fastica, fit_ebproj$vbar)
)
if (!intercept_degenerate) {
  scores[["FastICA + intercept"]] <- orient_and_scale(
    score_fastica_intercept,
    fit_ebproj$vbar
  )
}

round(sapply(scores, function(x) tapply(x, cell_type, mean)), 2)
round(cor(do.call(cbind, scores)), 2)

label_R2 <- sapply(
  scores,
  function(x) summary(lm(x ~ cell_type))$r.squared
)
round(label_R2, 3)

old_par <- par(mfrow = c(2, 2), mar = c(8, 4, 3, 1))
for (method in names(scores)) {
  boxplot(
    scores[[method]] ~ cell_type,
    main = method,
    xlab = "",
    ylab = "Standardized score",
    las = 2
  )
}
par(old_par)
```

As in the equal-mixture experiment, a single rank-one component can recover
only one split among the four known populations. The unequal mixture tests
whether the learned mixture weight and shift adapt to class imbalance.

The filtered dataset contains 5,079 cells, including 1,600 CD14 monocytes.
Both EB methods isolate CD14 monocytes from the other three populations and
produce identical scores, with \(\rho=0.980\) and label \(R^2=0.976\). The
fitted prior weight is \(\pi=0.310\), close to the observed CD14 proportion
\(1600/5079=0.315\). Ordinary FastICA finds a somewhat different contrast
(\(R^2=0.950\), correlation 0.72 with EB), whereas FastICA with an intercept
closely matches EB (\(R^2=0.969\), correlation 0.99) and does not degenerate.
