---
title: "Phase transition for fastICA"
author: "junmingguan"
date: "2026-06-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

## Introduction

The vanilla fastICA assumes a noiseless model, or at best a model with fixed noise level $\sigma^2=1$. In this analysis, we investigate what the smallest signal-to-noise ratio (SNR) above which fastICA or EBICA can reliably recover the true sources.

## Related works

Relevant references include @Li2022-lx, @Li2023-pc and @CelentanoTapZ2. These papers study AMP and related algorithms for spiked Wigner models. A canonical model in @Li2022-lx is

$$
M = \lambda vv^\top + W, \quad W_{ij}\sim N(0, 1/n), \quad W_{ii}\sim N(0, 2/n), \quad v_{i}\sim \text{Unif}(\pm 1/\sqrt{n}).
$$
Here $\lambda$ is the SNR parameter. The phase transition is at $\lambda=1$: below this threshold, the Bayes-optimal estimate has asymptotically zero correlation with the spike, while above it nontrivial correlation is achievable. The finite-sample AMP result in @Li2022-lx assumes a slightly stronger condition $\lambda > 1 + \log(n)n^{-1/16}$, but this extra gap is a proof condition rather than the sharp asymptotic threshold.

The simulation here starts from a noisy rank-$K$ model,

$$
Y = LF^\top + E, \quad E_{ij}\sim N(0,\sigma^2),
$$
with $L_{ik}\in\{\pm 1\}$ and $F_{jk}\sim N(0,c)$. Conditional on $L$,

$$
\mathbb{E}\left[\frac1p YY^\top \;\middle|\; L\right]=\sigma^2 I_n+c LL^\top.
$$

Thus the model induces a spiked covariance. The Li--Wei threshold does not directly apply to our simulation model, but Appendix A gives a heuristic reduction from the off-diagonal Gram matrix to a spiked Wigner model. Under this reduction, the effective Wigner spike strength is approximately

$$
\lambda_{\mathrm{eff}}
\approx
\frac{\lambda_{\mathrm{signal}}\sqrt{p/n}}
{\sqrt{\sigma^4+2\bar c\sigma^2}},
$$
where $\lambda_{\mathrm{signal}}$ is a signal eigenvalue of $(LF^\top)(LF^\top)^\top/p$ and

$$
\bar c
=
\frac1n \operatorname{tr}\left((LF^\top)(LF^\top)^\top/p\right)
$$

is the average signal variance per row. For recovering any one of the true sources, a conservative scale is given by the weakest nonzero signal eigenvalue, although the rectangular covariance model has its own BBP scaling and this mapping should be viewed as a heuristic connection.

With the phase-transition threshold $\lambda_{\mathrm{eff}}=1$, the corresponding phase-transition noise variance is approximately

$$
\sigma^2_{\mathrm{crit}}
\approx
\sqrt{\bar c^2+\lambda_{\mathrm{signal}}^2p/n}
-\bar c.
$$


## Results


We look at the unbalanced group example with varying noise levels, and apply rank-1 EBCD with Rademacher prior and fixed $\tau=1$, which is equivalent to fastICA and EBICA.

```{r}
source(file.path("code", "utils.R"))
source(file.path("code", "ebnm_shifted_rademacher.R"))
source(file.path("code", "ebnm_rademacher.R"))
source(file.path("code", "ebcd_functions.R"))
source(file.path("code", "ica_functions.R"))
```

```{r}
ebcd_rank_one_random_init <- function(X, init_Z = NULL, 
                                      tau = NULL, est_precision = TRUE,
                                      tol_backfit = 1e-6,
                                      maxiter_backfit = 100) {
  ebcd_obj <- ebcd_random_init(X = X, Z = init_Z, tau = tau, ebnm_fn = ebnm_rademacher) |>
    ebcd_backfit(
      tol = tol_backfit,
      maxiter = maxiter_backfit,
      est_precision = est_precision
    )
  return(ebcd_obj)
}
```

```{r}
K=9
p = 1000
n = 100
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,20),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

# LFt = L %*% t(FF)
```

```{r, include=FALSE}
# X1 = preprocess(X, n.comp = K)
# X1 = rbind(X1, rep(1, n))
```

```{r, include=FALSE}
# init_Z = rnorm(dim(X1)[1])
# ebcd.fit = ebcd_rank_one_random_init(X1, init_Z = init_Z, tau = 1, est_precision = FALSE, maxiter_backfit = 100)
# cor(ebcd.fit$EL, L)
# note that this is equivalent to ebica with s=d^-0.25
# ebica.res <- ebica(
#   X1,
#   W_init = init_Z,
#   symmetric_rademacher = TRUE,
#   s = nrow(X1)^(-1/4),
#   max_iter = 100,
#   conv_crit = "maxiter"
# )
# cor(ebica.res$S_plus, L)
```

We vary the variance of the Gaussian noise added to $LF^\top$, and count how many out of 100 random starts recover one of the true sources. Note that $\sigma^2_{\mathrm{crit}}\approx 125.742$.

```{r}
n_starts <- 100
signal <- L %*% t(FF)
signal_eigenvalues <- eigen(signal %*% t(signal) / p, symmetric = TRUE)$values
weakest_signal_eigenvalue <- signal_eigenvalues[K]
signal_variance_per_row <- sum(signal_eigenvalues) / n
critical_noise_variance <- sqrt(
  signal_variance_per_row^2 + weakest_signal_eigenvalue^2 * p / n
) - signal_variance_per_row

critical_noise_variance
```



```{r}

effective_lambdas <- c(3, 2, 1.2, 1, 0.75, 0.5)
noise_variances <- critical_noise_variance / effective_lambdas

run_ebcd_noise_experiment <- function(noise_var, noise_id) {
  set.seed(1000 + noise_id)
  X_noise <- signal + rnorm(n * p, mean = 0, sd = sqrt(noise_var))
  X1_noise <- preprocess(X_noise, n.comp = K)
  X1_noise <- rbind(X1_noise, rep(1, ncol(X1_noise)))

  max_abs_cor <- numeric(n_starts)

  for (start_id in 1:n_starts) {
    set.seed(start_id)
    init_Z <- rnorm(nrow(X1_noise))
    ebcd.fit <- ebcd_rank_one_random_init(
      X1_noise,
      init_Z = init_Z,
      tau = 1,
      est_precision = FALSE,
      maxiter_backfit = 100
    )
    max_abs_cor[start_id] <- max(abs(cor(ebcd.fit$EL, L)))
  }

  data.frame(
    noise_variance = noise_var,
    n_success = sum(max_abs_cor > 0.9),
    n_starts = n_starts,
    success_rate = mean(max_abs_cor > 0.9),
    mean_max_abs_cor = mean(max_abs_cor),
    max_max_abs_cor = max(max_abs_cor)
  )
}

ebcd_noise_results <- do.call(
  rbind,
  Map(run_ebcd_noise_experiment, noise_variances, seq_along(noise_variances))
)

ebcd_noise_results
```


## Appendix A: Heuristic Wigner reduction

This derivation is heuristic. It maps the off-diagonal part of the rectangular Gram matrix to the spiked Wigner scaling in @Li2022-lx.

Write

$$
S=LF^\top,
\quad
Y=S+E.
$$

Then

$$
\frac1pYY^\top
=
\frac1p SS^\top
+ \frac1p S E^\top
+ \frac1p E S^\top
+ \frac1p EE^\top.
$$

Let $\theta$ be an eigenvalue of $SS^\top/p$ with unit eigenvector $q$. The corresponding signal term is

$$
\theta qq^\top.
$$

For $i\neq j$, the three off-diagonal noise terms are

$$
\frac1p\sum_{\ell=1}^p S_{i\ell}E_{j\ell},
\quad
\frac1p\sum_{\ell=1}^p E_{i\ell}S_{j\ell},
\quad
\frac1p\sum_{\ell=1}^p E_{i\ell}E_{j\ell}.
$$

They have mean zero. Conditional on $S$, their variances are

$$
\frac{\sigma^2\|S_{i\cdot}\|^2}{p^2},
\quad
\frac{\sigma^2\|S_{j\cdot}\|^2}{p^2},
\quad
\frac{\sigma^4}{p}.
$$

Using the average row energy

$$
\bar c
=
\frac1n\sum_{i=1}^n\frac{\|S_{i\cdot}\|^2}{p}
=
\frac1n \operatorname{tr}(SS^\top/p),
$$

the typical off-diagonal Gram noise variance is approximately

$$
\frac{\sigma^4+2\bar c\sigma^2}{p}.
$$

Multiplying the centered off-diagonal Gram matrix by

$$
\alpha
\approx
\sqrt{\frac{p}{n(\sigma^4+2\bar c\sigma^2)}}
$$

puts the noise variance on the Wigner scale $1/n$. Under the same rescaling, the signal becomes approximately $\alpha\theta qq^\top$, so

$$
\lambda_{\mathrm{eff}}
\approx
\alpha\theta
\approx
\frac{\theta\sqrt{p/n}}{\sqrt{\sigma^4+2\bar c\sigma^2}}.
$$

Setting $\lambda_{\mathrm{eff}}\approx 1$ with $\theta=\lambda_{\mathrm{signal}}$ gives

$$
\sigma^4 + 2\bar c\sigma^2
\approx
\lambda_{\mathrm{signal}}^2 p/n,
$$

and therefore

$$
\sigma^2_{\mathrm{crit}}
\approx
\sqrt{\bar c^2+\lambda_{\mathrm{signal}}^2p/n}
-\bar c.
$$

In the rank-one special case, $\bar c\approx c$ and $\lambda_{\mathrm{signal}}\approx cn$, so this reduces to

$$
\sigma^2_{\mathrm{crit}}
\approx
c\left(\sqrt{1+np}-1\right).
$$





## References
