---
title: "On Rank-1 flashier update with Normal prior + Rademacher prior on whitened data"
author: "junmingguan"
date: "2026-05-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---



Here we demonstrate that flashier update with Normal and Rademacher prior might eventually collapse to 0. 

## Derivation


The model is

$$
Y \in \mathbb{R}^{p\times n} = vz^\top + E ,
$$

with a Rademacher prior on $v$, a normal prior on $z$, and an estimated noise variance:

$$
v_i \sim \operatorname{Rad}(1/2), \qquad
z_j \sim N(0,c\sigma^2), \qquad
E_{ij}\sim N(0,\sigma^2).
$$

Denote

$$
S=YY^\top, \qquad
\bar{v}=E_q[v], \qquad
\lambda=\frac{\bar{v}^\top S \bar{v}}{\overline{v^\top v}} = \frac{\bar{v}^\top S \bar{v}}{n}.
$$

Given $q_v$, and $\sigma^2$, the shrinkage factor for the posterior of $z$ is

$$
\alpha = \left(1-\frac{p \sigma^2}{\lambda}\right)_+,
$$

hence

$$
\bar{z} = \alpha Y^\top \bar{v}/n
$$

and

$$
\overline{z^\top z} = \alpha \lambda / n.
$$

Plugging in $q_z$ and $\sigma^2$, the update for $q_v$ is

$$
\bar{v}=\tanh\left(\frac{\alpha}{n \sigma^2}S \bar{v}\right)
$$


The noise variance update is

$$
\hat{\sigma}^2 = \frac{\operatorname{tr}(S) - \hat{\alpha} \lambda}{np}.
$$


- If $\hat{\alpha} = 0$, $\hat{\sigma}^2 = \operatorname{tr}(S)/np$
, $\bar{v}=\tanh 0 = 0$; this happens when $\lambda < \operatorname{tr}(S)/n=p$ for whitened $Y$. 
- If $\hat{\alpha} > 0$, we have $\widehat\sigma^2=\frac{\operatorname{tr}(S)-\lambda}{p(n-1)}$.  this happens when $\lambda \ge \operatorname{tr}(S)/n=p$. 

Let's assume we start with $\bar{v}$ such that $\lambda>p$. Using estimates of $\alpha$ and $\sigma^2$,

$$
\frac{\hat\alpha}{\hat\sigma^2}
=
\frac{np(\lambda-p)}{\lambda(np-\lambda)}.
$$

Define

$$
\gamma(\lambda)
=
\frac{np(\lambda-p)}{\lambda(np-\lambda)}.
$$

Then the update is

$$
 \bar{v}^+=
 \tanh\left(\gamma(\lambda)P\bar{v}\right),
$$

where $P=S/n$ is a projection matrix, assuming that $Y$ is whitened, and $\lambda=\|P\bar{v}\|^2$.

The next estimate $\bar{v}^+$ satisfies

$$
\lambda^+
\le
\left\|\tanh(\gamma P\bar{v})\right\|^2
\le \|P\bar{v}\|^2 = \lambda,
$$

whenever, $p<\lambda\leq p \sqrt{n}$. (could also hold for $\lambda > p\sqrt{n}$)

This tells us that $\lambda^+ \downarrow p$. But once $\lambda$ is sufficiently close to $p$, the next iteration will cross the threshold, yielding $\bar{v}^+ = 0$.


## Example

```{r}
library(flashier)
code_dir <- if (dir.exists("code")) "code" else "../code"
source(file.path(code_dir, "ica_functions.R"))
source(file.path(code_dir, "ebnm_rademacher.R"))


flash_x_rademacher <- function(X, S_init, ebnm_fn = ebnm_normal, S = NULL, var_type = 0, backfit = TRUE, maxiter = 30) {
   if (is.vector(S_init)) S_init <- matrix(S_init, ncol = 1)
  A_init = t(solve(crossprod(S_init), crossprod(S_init, t(X))))
  if (!is.null(S)) {
    var_type = NULL
  }
  lambda_path = c()
  res <- flash_init(X, var_type = var_type, S = S) |>
      flash_factors_init(
        init = list(A_init, S_init),
        ebnm_fn = list(ebnm_fn, ebnm_rademacher)
  )
  lambda_path[1] = t(S_init) %*% t(X) %*% X %*% S_init
  for (i in 1:maxiter) {
    res <- res |>
      flash_backfit(verbose = 0, extrapolate = TRUE, maxiter = 1)
    lambda_path[i+1] = t(res$F_pm) %*% t(X) %*% X %*% res$F_pm
  }
  return(list(res=res, lambda_path = lambda_path))
}
```
```{r}
K=3
p = 100
n = 1000
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,20),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

X = L %*% t(FF) + rnorm(n*p,0,0.01)
plot(X %*% FF[,1])

X1 = preprocess(X, n.comp = 9)
X1 = rbind(rep(1,ncol(X1)), X1)
```

```{r}
#| warning: false

fit = flash_x_rademacher(X1, L[,1])

```

We see here that the fit is essentially 0, and $\lambda$ converges to 0, although initially it is larger than $p=100$.

```{r}
sum(abs(fit$res$F_pm))
fit$lambda_path / n
```
