---
title: "On Rank-1 flashier update with Normal prior + Rademacher prior on whitened data"
author: "junmingguan"
date: "2026-05-22"
output: workflowr::wflow_html
editor_options:
 chunk_output_type: console
---

<style>
.theorem,
.proof {
  border-left: 4px solid #e95420;
  margin: 1.25em 0;
  padding: 0.85em 1em;
  background: #f8f8f8;
}

.proof {
  border-left-color: #666;
}

.theorem::before,
.proof::before {
  display: block;
  font-weight: 700;
  margin-bottom: 0.35em;
}

.theorem::before {
  content: "Theorem.";
}

.proof::before {
  content: "Proof.";
}
</style>



Here we demonstrate that flashier update with Normal and Rademacher prior might eventually collapse to 0. 

## Derivation


The model is

$$
Y \in \mathbb{R}^{n\times p} = vz^\top + E ,
$$

with a Rademacher prior on $v$, a normal prior on $z$, and an estimated noise variance:

$$
v_i \sim \operatorname{Rad}(1/2), \qquad
z_j \sim N(0,c\sigma^2), \qquad
E_{ij}\sim N(0,\sigma^2).
$$

Denote

$$
S=YY^\top, \qquad
P=S/n, \qquad
\bar{v}=E_q[v], \qquad
\lambda=\frac{\bar{v}^\top S \bar{v}}{\overline{v^\top v}} = \frac{\bar{v}^\top S \bar{v}}{n}.
$$
where we assume $Y$ is whitened, so that $P$ is a projection matrix.


Given $q_v$, and $\sigma^2$, the shrinkage factor for the posterior of $z$ is

$$
\boxed{\alpha = \left(1-\frac{p \sigma^2}{\lambda}\right)_+}
$$

Hence

$$
\boxed{\bar{z} = \alpha Y^\top \bar{v}/n}
$$

and

$$
\boxed{\overline{z^\top z} = \alpha \lambda / n}
$$

Plugging in $q_z$ and $\sigma^2$, the update for $q_v$ is

$$
\boxed{\bar{v}=\tanh\left(\frac{\alpha}{n \sigma^2}S \bar{v}\right) = \tanh\left(\frac{\alpha}{\sigma^2}P \bar{v}\right)}
$$

### Understanding ELBO with respect to $c$

The ELBO for fixed $c$ and $\sigma^2$ is given by

$$
\boxed{
\begin{aligned}
\text{ELBO}
\text{ }=&
-\frac{np}{2}\log(2\pi\sigma^2)
-\frac{1}{2\sigma^2}
\left[
\|Y\|_F^2
-2\bar v^\top Y\bar z
+n\overline{z^\top z}
\right]
\\
& \text{ }-
\sum_{i=1}^n
\left[
\frac{1+\bar v_i}{2}\log(1+\bar v_i)
+
\frac{1-\bar v_i}{2}\log(1-\bar v_i)
\right]
\\
& \text{ } -
\frac{1}{2}
\left[
\frac{\overline{z^\top z}}{c\sigma^2}
-p + p\log(c\sigma^2)
- \sum_{j=1}^p\log \underbrace{\left(\overline{z_j^2} - \bar{z}_j^2\right)}_{=:\eta^2=\operatorname{Var}_q(z_j)}
\right]
\end{aligned}
}
$$
Maximizing the ELBO over $c$ with others fixed gives

$$
\boxed{
\hat{c}
=
\frac{\overline{z^\top z}}{p\sigma^2}
}
$$

Plugging in $c$ into $q_z$, we have

$$
\boxed{
\bar{z} = \frac{c}{1+nc}Y^\top \bar{v}
\qquad
\eta_j^2 = \frac{c\sigma^2}{1+nc}
}
$$
So the ELBO profiled over $z$ with $\sigma^2$ fixed is 

$$
\begin{aligned}
\widehat{\text{ELBO}}(c, q_v)
=\text{ }&
\text{const.} + \frac{c}{1+nc}
\frac{\|Y^\top \bar v\|^2}{2\sigma^2}
\\
&-
\sum_{i=1}^n
\left[
\frac{1+\bar{v}_i}{2}\log(1+\bar{v}_i)
+
\frac{1-\bar{v}_i}{2}\log(1-\bar{v}_i)
\right]
\\
&
-\frac p2\log(1+nc).
\end{aligned}
$$
Then

$$
\hat{c} = 
\frac{1}{n}\left(\frac{\|Y^\top \bar{v}\|^2}{pn\sigma^2} - 1\right)_+
=
\frac{1}{n}\left(\frac{\|Y^\top \bar{v}\|^2}{pn\sigma^2} - 1\right)_+.
$$

So whether the ELBO is maximized at $c=0$ for fixed $q_v$ depends on whether $\|Y^\top\bar v\|^2 = n\|P\bar{v}\|^2 > pn \sigma^2$.

But if we look at $c$ and $q_v$ jointly: since $\text{KL}(q_v \| \text{Rad}(1/2)) \ge \frac12 \|\bar{v}\|^2$ (equality achieved when $\bar v=0$)

$$
\begin{aligned}
\widehat{\text{ELBO}}(c, q_v) 
\leq \text{ } & 
\text{const.} +\frac{c}{1+nc}  \frac{\|Y^\top \bar v\|^2}{2\sigma^2}
\\
& + \frac12 \|\bar v\|^2 -\frac{p}{2}\log (1+nc)
\\
\leq \text{ } & 
\text{const.} +\frac{c}{1+nc}  \frac{n \|\bar v\|^2}{2\sigma^2}
\\
& + \frac12 \|\bar v\|^2 -\frac{p}{2}\log (1+nc) \\
= \text{ } & 
\text{const.} 
-\frac p2\log(1+nc) +
\frac12
\left[
\frac{nc}{\sigma^2(1+nc)}-1
\right]
\|\bar v\|^2=:U(c, \bar v).
\end{aligned}
$$
For $\sigma^2\ge 1$, $\frac{nc}{\sigma^2(1+nc)}<1$. Ignoring the constant, $U(c, \bar v)\le 0$ and is maximized at $c=0$ and $\bar{v}=0$, hence the profiled $\widehat{\text{ELBO}}(c, q_v)$ is also maximized at $c=0$ and $\bar{v}=0$ and the maximum is $0$.

### $\sigma^2$ not fixed

The noise variance update is

$$
\boxed{\hat{\sigma}^2 = \frac{\operatorname{tr}(S) - \hat{\alpha} \lambda}{np}}
$$


- If $\hat{\alpha} = 0$, $\hat{\sigma}^2 = \operatorname{tr}(S)/np$
, $\bar{v}=\tanh 0 = 0$; this happens when $\lambda < \operatorname{tr}(S)/n=p$ for whitened $Y$. 
- If $\hat{\alpha} > 0$, we have $\hat\sigma^2=\frac{\operatorname{tr}(S)-\lambda}{p(n-1)}$.  this happens when $\lambda \ge \operatorname{tr}(S)/n=p$. 

::: {.theorem #thm:estimated-sigma-collapse}
Suppose we start with an initialization $\bar{v}_0\in\{\pm1\}^n$ such that $p<\lambda_0<p\sqrt{n}$. Then for some finite $t$ we have $\bar{v}_t=0$.
:::

::: {.proof}
Using estimates of $\alpha$ and $\sigma^2$, for the $t$th step,

$$
\frac{\hat\alpha}{\hat\sigma^2}
=
\frac{np(\lambda_t-p)}{\lambda_t(np-\lambda_t)}.
$$

Define

$$
\gamma(\lambda_t)
=
\frac{np(\lambda_t-p)}{\lambda_t(np-\lambda_t)}.
$$

Then the update is

$$
 \bar{v}_{t+1}=
 \tanh\left(\gamma(\lambda_t)P\bar{v}_t\right),
$$


The next estimate $\bar{v}_{t+1}$ satisfies

$$
\lambda_{t+1} = \|P\bar{v}_{t+1}\|^2 = \|P\tanh(\gamma(\lambda_t) P\bar{v}_t)\|^2
\le
\left\|\tanh(\gamma(\lambda_t) P\bar{v}_{t})\right\|^2
\le \|P\bar{v}_t\|^2 = \lambda_t,
$$

whenever, $p<\lambda_t\leq p \sqrt{n}$, so that $\gamma(\lambda_t) \le 1$. (might also hold for $\lambda_t > p\sqrt{n}$)

This tells us that $\lambda_{t} \searrow p$. But once $\lambda_{t}$ is sufficiently close to $p$, the next iteration will cross the threshold, yielding $\bar{v}^{t+1} = 0$. TODO: more rigorous statement?
:::


### Fixed $\sigma^2$

For fixed $\sigma^2$, a factor will be removed if $\lambda < p\cdot \sigma^2$. However, since the update is (suppose $\lambda < p \sigma^2$, otherwise the update is 0)

$$
\bar{v}^+ = \tanh(\frac{\alpha}{ \sigma^2}S \bar{v}/n) = \tanh\left(\left(\frac{1}{\sigma^2}−\frac{p}{\lambda}\right) \frac{1}{n}S \bar{v}\right)
$$

we want $\left(\frac{1}{\sigma^2}−\frac{p}{\lambda}\right)>1$, or equivalently $\lambda> \frac{p\sigma^2}{1-\sigma^2}$ so that the update is not contrasive. 

But since $Y$ is whitened, $\lambda\le n$. So $\frac{1}{\sigma^2}−\frac{p}{\lambda} \le \frac{1}{\sigma^2}-\frac{p}{n}$, and the right hand size is less than 1 if $\sigma^2>\frac{n}{n+p}$.

In other words, to ensure the update doesn't collapse to 0, we need both $\sigma^2<\frac{n}{n+p}$ and an initialization with $\lambda < p \sigma^2$.

::: {.theorem #thm:fixed-sigma-collapse}
If $\sigma^2>\frac{n}{n+p}$, then for any initialization $\bar{v}_0\in\{\pm 1\}^n$, we have $\lambda_t<p\sigma^2$ for some $t$, hence $\bar{v}_{t+1}=0$.
:::

::: {.proof}
Suppose the initialization $\bar{v}_0$ satisfies $\lambda_0>p\sigma^2$, and that $n \le p\sigma^2$ (otherwise it's trivial). Let

$$
\rho:= \frac{1}{\sigma^2}-\frac{p}{n}.
$$

Denote

$$
\gamma(\lambda_t) = \frac{1}{\sigma^2}-\frac{p}{\lambda_t},
$$

so $\gamma(\lambda)<1$. Since $P:=S/n$ is a projection matrix, we have

$$
\lambda_{t+1} = \|P \bar{v}_{t+1}\|^2 = \|P\tanh(\gamma(\lambda_t) P \bar{v}_t)\|^2 \le \|\tanh(\gamma(\lambda_t) P \bar{v}_t)\|^2 \le \gamma(\lambda_{t})^2\| P \bar{v}_t\|^2.
$$

So

$$
\lambda_{t+1} \le \rho^2 \lambda_t,
$$

and

$$
\lambda_t \leq \rho^{2t} \lambda_0.
$$

Hence there exists $t$ such that

$$
\lambda_t \le p \sigma^2.
$$
:::

## Example

```{r}
library(flashier)
library(pheatmap)
code_dir <- if (dir.exists("code")) "code" else "../code"
source(file.path(code_dir, "ica_functions.R"))
source(file.path(code_dir, "ebnm_rademacher.R"))
source(file.path(code_dir, "ebnm_point_rademacher.R"))
source(file.path(code_dir, "ebnm_normal_fixed_sd.R"))
source(file.path(code_dir, "utils.R"))


flash_x_rademacher <- function(X, S_init, ebnm_fn = ebnm_normal, point_mass = FALSE, S = NULL, var_type = 0, backfit = TRUE, maxiter = 30) {
   if (is.vector(S_init)) S_init <- matrix(S_init, ncol = 1)
  A_init = t(solve(crossprod(S_init), crossprod(S_init, t(X))))
  if (!is.null(S)) {
    var_type = NULL
  }
  lambda_path = c()
  elbo_path <- c()
  resid_path <- c()
  z_prior_var_path <- c()
  if (point_mass) {
    ebnm_source = ebnm_point_generalized_rademacher
  }
  else {
    ebnm_source = ebnm_rademacher

  res <- flash_init(X, var_type = var_type, S = S) |>
      flash_factors_init(
        init = list(A_init, S_init),
        ebnm_fn = list(ebnm_fn, ebnm_source)
  )
  lambda_path[1] = t(S_init) %*% t(X) %*% X %*% S_init
  elbo_path[1] = res$elbo
  resid_path[1] = res$residuals_sd
  z_prior_var_path[1] = res$L_ghat[[1]]$sd
  for (i in 1:maxiter) {
    res <- res |>
      flash_backfit(verbose = 0, extrapolate = TRUE, maxiter = 1)
    lambda_path[i+1] = t(res$F_pm) %*% t(X) %*% X %*% res$F_pm
    elbo_path[i+1] = res$elbo
    resid_path[i+1] = res$residuals_sd
    z_prior_var_path[i+1] = res$L_ghat[[1]]$sd
  }
  return(list(res=res, lambda_path = lambda_path, 
              elbo_path = elbo_path, resid_path = resid_path,
              z_prior_var_path = z_prior_var_path))
  }
}
```

```{r}
K=9
p = 1000
n = 100
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,20),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

X = L %*% t(FF) + rnorm(n*p,0,0.01)
plot(X %*% FF[,1])

X1 = preprocess(X, n.comp = 9)
X1 = rbind(rep(1,ncol(X1)), X1)
```

```{r, warning = FALSE}
#| warning: false

fit_fixed = flash_x_rademacher(X1, L[,1], S=1, ebnm_fn = ebnm_normal)
fit_notfixed = flash_x_rademacher(X1, L[,1], S=NULL, ebnm_fn = ebnm_normal)

```

We see here that the fit (either with $\sigma^2$ fixed or estimated) is essentially 0, and $\lambda$ converges to 0, although initially it is larger than $p=10$ ($1+K=10$). Also, the estimated prior variance for $z$ converges to 0, and the ELBO seems to stablized at $c=0$.

```{r}
sum(abs(fit_fixed$res$F_pm))

sum(abs(fit_notfixed$res$F_pm))

plot(fit_fixed$z_prior_var_path[-1]^2, xlab = 'Iteration', ylab = 'Estimated prior variance')
plot(fit_fixed$elbo_path[-1], xlab = 'Iteration', ylab = 'ELBO')
plot(fit_fixed$lambda_path / n, xlab = 'Iteration', ylab = 'lambda')
plot(fit_notfixed$z_prior_var_path[-1]^2, xlab = 'Iteration', ylab = 'Estimated prior variance')
plot(fit_notfixed$elbo_path[-1], xlab = 'Iteration', ylab = 'ELBO')
plot(fit_notfixed$lambda_path / n, xlab = 'Iteration', ylab = 'lambda')
```


### Unconstrained $z$

Here, we examine the case where $z# is unconstrained and estimated via MLE.

To do it in flashier, we can set and fix the normal prior variance to be very large. It seems that $z=0$ is still a fixed point for whitened data, but it takes a large number of iterations to converge to it.

```{r, warning = FALSE}
ebnm_normal_fixed_prior_sd <- make_ebnm_normal_fixed_sd(100000)
fit_fixed_prior_sd = flash_x_rademacher(X1, L[,1], S=NULL, ebnm_fn = ebnm_normal_fixed_prior_sd, maxiter = 5000)
sum(abs(fit_fixed_prior_sd$res$F_pm))
plot(fit_fixed_prior_sd$lambda_path / n, xlab = 'Iteration', ylab = 'lambda')
plot(fit_fixed_prior_sd$elbo_path[-1], xlab = 'Iteration', ylab = 'ELBO')
plot(fit_fixed_prior_sd$resid_path, xlab = 'Iteration', ylab = 'Residual variance')
```


A better way is to fit thr true MLE. Without constraint, convergence to $z=0$ is even slower, although the ELBO stablizes very quickly.

Is $z=0$ maximizing the ELBO? If seems that if we keep going regardless of the ELBO, we will reach $z=0$ eventually, but in practice we might have stopped because ELBO won't change much.
```{r}
ebica_fit = ebica_rank_one_unconstrained(X1, symmetric_rademacher = TRUE, max_iter = 50000)
max(apply(cor(L, t(X1) %*% ebica_fit$W), MARGIN = 1, abs))
sum(abs(ebica_fit$S_plus))
ebica_fit$S_plus
plot(ebica_fit$obj, xlab = 'Iteration', ylab = 'ELBO')
plot(ebica_fit$residual_variance_path, xlab = 'Iteration', ylab = 'Residual variance')
```