---
title: "On generalized Rademacher"
author: "Junming Guan"
date: "2026-06-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


<style>
.theorem,
.proof {
  border-left: 4px solid #e95420;
  margin: 1.25em 0;
  padding: 0.85em 1em;
  background: #f8f8f8;
}

.proof {
  border-left-color: #666;
}

.theorem::before,
.proof::before {
  display: block;
  font-weight: 700;
  margin-bottom: 0.35em;
}

.theorem::before {
  content: "Theorem.";
}

.proof::before {
  content: "Proof.";
}
</style>

## Introduction

In this analysis, I compare EBICA with a Rademacher prior, EBICA with a generalized Rademacher prior (where the prior probability $\pi$ is estimated by maximizing the marginal likelihood), and fastICA, on the whitened data with and without an intercept in the 9 unbalanced group examples.

Without an intercept, none of the three approaches appears to capture one true source; for EBICA generalized Rademacher, the estimated prior probability also seems far off from the true 0.2/0.8 probabilities.

With the intercept added, EBICA generalized Rademacher always picks up the trivial solution, with $\hat{\pi}=1/0$, which is not ideal; I wonder if it makes sense to put a prior on $\pi$ to shrink it toward $0.5$, say $\text{Beta}(5, 5)$? EBICA Rademacher, on the other hand, is very similar to fastICA as expected, but fastICA seems to capture the trivial solution slightly more often, probably due to the Newton "acceleration".

```{r}
library(flashier)
library(pheatmap)
code_dir <- if (dir.exists("code")) "code" else "../code"
source(file.path(code_dir, "ica_functions.R"))
source(file.path(code_dir, "ebnm_rademacher.R"))
source(file.path(code_dir, "ebnm_point_rademacher.R"))
source(file.path(code_dir, "ebnm_normal_fixed_sd.R"))
source(file.path(code_dir, "utils.R"))


flash_x_rademacher <- function(X, S_init, ebnm_fn = ebnm_normal, ebnm_source = ebnm_rademacher, S = NULL, var_type = 1, backfit = TRUE, maxiter = 30) {
   if (is.vector(S_init)) S_init <- matrix(S_init, ncol = 1)
  A_init = t(solve(crossprod(S_init), crossprod(S_init, t(X))))
  if (!is.null(S)) {
    var_type = NULL
  }
  lambda_path = c()
  elbo_path <- c()
  resid_path <- c()
  z_prior_var_path <- c()

  res <- flash_init(X, var_type = var_type, S = S) |>
      flash_factors_init(
        init = list(A_init, S_init),
        ebnm_fn = list(ebnm_fn, ebnm_source)
  )
  lambda_path[1] = t(S_init) %*% t(X) %*% X %*% S_init
  elbo_path[1] = res$elbo
  resid_path[1] = res$residuals_sd
  z_prior_var_path[1] = res$L_ghat[[1]]$sd
  for (i in 1:maxiter) {
    res <- res |>
      flash_backfit(verbose = 0, extrapolate = TRUE, maxiter = 1)
    lambda_path[i+1] = t(res$F_pm) %*% t(X) %*% X %*% res$F_pm
    elbo_path[i+1] = res$elbo
    resid_path[i+1] = res$residuals_sd
    z_prior_var_path[i+1] = res$L_ghat[[1]]$sd
  }
  return(list(res=res, lambda_path = lambda_path, 
              elbo_path = elbo_path, resid_path = resid_path,
              z_prior_var_path = z_prior_var_path))
}
```


```{r}
ebnm_rademacher_0.2 <- function(
    x,
    s = 1,
    mode = 0,
    g_init = NULL,
    fix_g = FALSE,
    output = NULL,
    control = NULL
) {
  ebnm_generalized_rademacher(
    x = x,
    s = s,
    mode = mode,
    m = 1,
    g_init = rademacher(pi = 0.2, m = 1),
    fix_g = TRUE
  )
}
```

## 9 unbalanced groups

```{r}
K=9
p = 1000
n = 100
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,20),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

X = L %*% t(FF) + rnorm(n*p,0,0.01)
plot(X %*% FF[,1])

X1 = preprocess(X, n.comp = K)
# X1 = rbind(rep(1,ncol(X1)), X1)
```

### No intercept

```{r}
generalized.rad.corr.max <- c()
rad.corr.max <- c()
rad.0.2.corr.max <- c()
fastica.corr.max <- c()

generalized.rad.obj <- c()
rad.obj <- c()
rad.0.2.obj <- c()
fastica.obj <- c()

generalized.rad.Lhat <- matrix(0, nrow = 100, ncol = 100)
rad.Lhat <- matrix(0, nrow = 100, ncol = 100)
rad.0.2.Lhat <- matrix(0, nrow = 100, ncol = 100)
fastica.Lhat <- matrix(0, nrow = 100, ncol = 100)

generalized.rad.prior.prob <- c()

for (seed in 1:100) {
  set.seed(seed = seed)
  W_init = rnorm(dim(X1)[1]) #X1 %*% L[,1]
  res.generalized.rad = ebica(X1, s = 1, W_init = W_init, max_iter = 100)
  res.rad = ebica(X1, s = 1, W_init = W_init, symmetric_rademacher = T, max_iter = 100)
  res.0.2.rad = ebica_generalized_parallel(X1, K=1, ebnm_fn = ebnm_rademacher_0.2, s = 1, W_init = W_init, max_iter = 100)
  fastica_rank_one_update(X1, init_w = W_init)
  
  generalized.rad.corr.max[seed] = max(abs(cor(L, res.generalized.rad$S)))
  generalized.rad.obj[seed] = compute_objective(X1, res.generalized.rad$W)
  generalized.rad.Lhat[seed,] = res.generalized.rad$S
  
  generalized.rad.prior.prob[seed] = ifelse(res.generalized.rad$prior_probs > 0.5, res.generalized.rad$prior_probs, 1 - res.generalized.rad$prior_probs)
  
  rad.corr.max[seed] = max(abs(cor(L, res.rad$S)))
  rad.obj[seed] = compute_objective(X1, res.rad$W)
  rad.Lhat[seed,] = res.rad$S
  
  rad.0.2.corr.max[seed] = max(abs(cor(L, res.0.2.rad$S)))
  rad.0.2.obj[seed] = compute_objective(X1, res.0.2.rad$W)
  rad.0.2.Lhat[seed,] = res.0.2.rad$S
}
```

FastICA

```{r}
init_W = matrix(0, ncol = 100, nrow = dim(X1)[1])

for (seed in 1:100) {
  set.seed(seed)
  init_W[,seed] = rnorm(dim(X1)[1])
}
W = fastica_vectorized_rank_one_update(X1, init_W = init_W, include_G2 = TRUE)

fastica.obj = apply(W, 2, FUN = function(w) {compute_objective(X1, w)})
fastica.corr.max = apply(W, 2, FUN = function(w) {max(abs(cor(L, t(X1) %*% w)))})
fastica.Lhat = t(t(X1) %*% W)

```


```{r}
plot(sort(generalized.rad.obj), ylab = "logcosh objective",
     main = 'Generalized Rademacher')
plot(generalized.rad.corr.max[order(generalized.rad.obj)], ylab = "Max correlation with a true source",
     main = 'Generalized Rademacher')
plot(generalized.rad.prior.prob[order(generalized.rad.obj)], ylab = "Estimated prior probability",
     main = 'Generalized Rademacher')
plot_heatmap(t(generalized.rad.Lhat[order(generalized.rad.obj), ]))


plot(sort(rad.obj), ylab = "logcosh objective",
     main = 'Rademacher')
plot(rad.corr.max[order(rad.obj)], ylab = "Max correlation with a true source",
     main = 'Rademacher')
plot_heatmap(t(rad.Lhat[order(rad.obj), ]))

plot(sort(rad.0.2.obj), ylab = "logcosh objective",
     main = 'Generalized Rademacher p=0.2')
plot(rad.0.2.corr.max[order(rad.0.2.obj)], ylab = "Max correlation with a true source",
     main = 'Generalized Rademacher p=0.2')
plot_heatmap(t(rad.0.2.Lhat[order(rad.0.2.obj), ]))

plot(sort(fastica.obj), ylab = "logcosh objective",
     main = 'fastICA')
plot(fastica.corr.max[order(fastica.obj)], ylab = "Max correlation with a true source",
     main = 'fastICA')
plot_heatmap(t(fastica.Lhat[order(fastica.obj), ]))
```




### With intercept

```{r}
X1 = preprocess(X, n.comp = K)
X1 = rbind(rep(1,ncol(X1)), X1)
```


```{r}
generalized.rad.corr.max <- c()
rad.corr.max <- c()
rad.0.2.corr.max <- c()
fastica.corr.max <- c()

generalized.rad.obj <- c()
rad.obj <- c()
rad.0.2.obj <- c()
fastica.obj <- c()

generalized.rad.Lhat <- matrix(0, nrow = 100, ncol = 100)
rad.Lhat <- matrix(0, nrow = 100, ncol = 100)
rad.0.2.Lhat <- matrix(0, nrow = 100, ncol = 100)
fastica.Lhat <- matrix(0, nrow = 100, ncol = 100)

generalized.rad.prior.prob <- c()

for (seed in 1:100) {
  set.seed(seed = seed)
  W_init = rnorm(dim(X1)[1]) #X1 %*% L[,1]
  res.generalized.rad = ebica(X1, s = 1, W_init = W_init, max_iter = 100)
  res.rad = ebica(X1, s = 1, W_init = W_init, symmetric_rademacher = T, max_iter = 100)
  res.0.2.rad = ebica_generalized_parallel(X1, K=1, ebnm_fn = ebnm_rademacher_0.2, s = 1, W_init = W_init, max_iter = 100)
  fastica_rank_one_update(X1, init_w = W_init)
  
  generalized.rad.corr.max[seed] = max(abs(cor(L, res.generalized.rad$S)))
  generalized.rad.obj[seed] = compute_objective(X1, res.generalized.rad$W)
  generalized.rad.Lhat[seed,] = res.generalized.rad$S
  
  generalized.rad.prior.prob[seed] = ifelse(res.generalized.rad$prior_probs > 0.5, res.generalized.rad$prior_probs, 1 - res.generalized.rad$prior_probs)
  
  rad.corr.max[seed] = max(abs(cor(L, res.rad$S)))
  rad.obj[seed] = compute_objective(X1, res.rad$W)
  rad.Lhat[seed,] = res.rad$S
  
  rad.0.2.corr.max[seed] = max(abs(cor(L, res.0.2.rad$S)))
  rad.0.2.obj[seed] = compute_objective(X1, res.0.2.rad$W)
  rad.0.2.Lhat[seed,] = res.0.2.rad$S
}
```

FastICA

```{r}
init_W = matrix(0, ncol = 100, nrow = dim(X1)[1])

for (seed in 1:100) {
  set.seed(seed)
  init_W[,seed] = rnorm(dim(X1)[1])
}
W = fastica_vectorized_rank_one_update(X1, init_W = init_W, include_G2 = TRUE)

fastica.obj = apply(W, 2, FUN = function(w) {compute_objective(X1, w)})
fastica.corr.max = apply(W, 2, FUN = function(w) {max(abs(cor(L, t(X1) %*% w)))})
fastica.Lhat = t(t(X1) %*% W)

```


```{r}
plot(sort(generalized.rad.obj), ylab = "logcosh objective",
     main = 'Generalized Rademacher')
plot(generalized.rad.corr.max[order(generalized.rad.obj)], ylab = "Max correlation with a true source",
     main = 'Generalized Rademacher')
plot(generalized.rad.prior.prob[order(generalized.rad.obj)], ylab = "Estimated prior probability",
     main = 'Generalized Rademacher')
plot_heatmap(t(generalized.rad.Lhat[order(generalized.rad.obj), ]))


plot(sort(rad.obj), ylab = "logcosh objective",
     main = 'Rademacher')
plot(rad.corr.max[order(rad.obj)], ylab = "Max correlation with a true source",
     main = 'Rademacher')
plot_heatmap(t(rad.Lhat[order(rad.obj), ]))

plot(sort(rad.0.2.obj), ylab = "logcosh objective",
     main = 'Generalized Rademacher p=0.2')
plot(rad.0.2.corr.max[order(rad.0.2.obj)], ylab = "Max correlation with a true source",
     main = 'Generalized Rademacher p=0.2')
plot_heatmap(t(rad.0.2.Lhat[order(rad.0.2.obj), ]))

plot(sort(fastica.obj), ylab = "logcosh objective",
     main = 'fastICA')
plot(fastica.corr.max[order(fastica.obj)], ylab = "Max correlation with a true source",
     main = 'fastICA')
plot_heatmap(t(fastica.Lhat[order(fastica.obj), ]))
```



## 9 unbalanced groups

Here I tried applying no-intercept generalized Rademacher with $\pi$ fixed to $0.2$, while the true $\pi$ is $0.15$.

```{r}
K=9
p = 1000
n = 100
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,10),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

X = L %*% t(FF) + rnorm(n*p,0,0.01)
plot(X %*% FF[,1])

X1 = preprocess(X, n.comp = K)
# X1 = rbind(rep(1,ncol(X1)), X1)
```


```{r}
generalized.rad.corr.max <- c()
rad.corr.max <- c()
rad.0.2.corr.max <- c()
fastica.corr.max <- c()

generalized.rad.obj <- c()
rad.obj <- c()
rad.0.2.obj <- c()
fastica.obj <- c()

generalized.rad.Lhat <- matrix(0, nrow = 100, ncol = 100)
rad.Lhat <- matrix(0, nrow = 100, ncol = 100)
rad.0.2.Lhat <- matrix(0, nrow = 100, ncol = 100)
fastica.Lhat <- matrix(0, nrow = 100, ncol = 100)

generalized.rad.prior.prob <- c()

for (seed in 1:100) {
  set.seed(seed = seed)
  W_init = rnorm(dim(X1)[1]) #X1 %*% L[,1]
  res.generalized.rad = ebica(X1, s = 1, W_init = W_init, max_iter = 100)
  res.rad = ebica(X1, s = 1, W_init = W_init, symmetric_rademacher = T, max_iter = 100)
  res.0.2.rad = ebica_generalized_parallel(X1, K=1, ebnm_fn = ebnm_rademacher_0.2, s = 1, W_init = W_init, max_iter = 100)
  fastica_rank_one_update(X1, init_w = W_init)
  
  generalized.rad.corr.max[seed] = max(abs(cor(L, res.generalized.rad$S)))
  generalized.rad.obj[seed] = compute_objective(X1, res.generalized.rad$W)
  generalized.rad.Lhat[seed,] = res.generalized.rad$S
  
  generalized.rad.prior.prob[seed] = ifelse(res.generalized.rad$prior_probs > 0.5, res.generalized.rad$prior_probs, 1 - res.generalized.rad$prior_probs)
  
  rad.corr.max[seed] = max(abs(cor(L, res.rad$S)))
  rad.obj[seed] = compute_objective(X1, res.rad$W)
  rad.Lhat[seed,] = res.rad$S
  
  rad.0.2.corr.max[seed] = max(abs(cor(L, res.0.2.rad$S)))
  rad.0.2.obj[seed] = compute_objective(X1, res.0.2.rad$W)
  rad.0.2.Lhat[seed,] = res.0.2.rad$S
}
```

FastICA

```{r}
init_W = matrix(0, ncol = 100, nrow = dim(X1)[1])

for (seed in 1:100) {
  set.seed(seed)
  init_W[,seed] = rnorm(dim(X1)[1])
}
W = fastica_vectorized_rank_one_update(X1, init_W = init_W, include_G2 = TRUE)

fastica.obj = apply(W, 2, FUN = function(w) {compute_objective(X1, w)})
fastica.corr.max = apply(W, 2, FUN = function(w) {max(abs(cor(L, t(X1) %*% w)))})
fastica.Lhat = t(t(X1) %*% W)

```


```{r}
plot(sort(generalized.rad.obj), ylab = "logcosh objective",
     main = 'Generalized Rademacher')
plot(generalized.rad.corr.max[order(generalized.rad.obj)], ylab = "Max correlation with a true source",
     main = 'Generalized Rademacher')
plot(generalized.rad.prior.prob[order(generalized.rad.obj)], ylab = "Estimated prior probability",
     main = 'Generalized Rademacher')
plot_heatmap(t(generalized.rad.Lhat[order(generalized.rad.obj), ]))


plot(sort(rad.obj), ylab = "logcosh objective",
     main = 'Rademacher')
plot(rad.corr.max[order(rad.obj)], ylab = "Max correlation with a true source",
     main = 'Rademacher')
plot_heatmap(t(rad.Lhat[order(rad.obj), ]))

plot(sort(rad.0.2.obj), ylab = "logcosh objective",
     main = 'Generalized Rademacher p=0.2')
plot(rad.0.2.corr.max[order(rad.0.2.obj)], ylab = "Max correlation with a true source",
     main = 'Generalized Rademacher p=0.2')
plot_heatmap(t(rad.0.2.Lhat[order(rad.0.2.obj), ]))

plot(sort(fastica.obj), ylab = "logcosh objective",
     main = 'fastICA')
plot(fastica.corr.max[order(fastica.obj)], ylab = "Max correlation with a true source",
     main = 'fastICA')
plot_heatmap(t(fastica.Lhat[order(fastica.obj), ]))
```

