---
title: "Analysis of cytokine-stimulated mouse islet cells"
author: "Junming Guan"
date: "2026-08-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Data and preprocessing

Here I analyze the pancreas cytokine data, retaining the cell-type and treatment-condition
labels for visualization. The goal is to investigate if our proposed methods can identify cell-type or condition contrasts. 

Genes expressed in fewer than 10 cells are removed,
and counts are library-size normalized before applying the shifted log transform.

```{r setup, message=FALSE}
library(irlba)
set.seed(1)
load("data/pancreas_cytokine_lsa.RData")
barcodes <- as.data.frame(barcodes)
sample_info <- data.frame(
  celltype = factor(barcodes$celltype,
                    c("Acinar", "Ductal", "Endothelial/Mesnchymal", "Macrophage",
                      "Alpha", "Beta", "Delta", "Gamma")),
  condition = factor(barcodes$condition,
                     c("Untreated", "IL-1B", "IFNg", "IL-1B_IFNg"))
)
n <- nrow(counts)
table(sample_info$celltype, sample_info$condition)
```

```{r preprocess}
# Remove genes that are expressed in fewer than 10 cells.
x <- colSums(counts > 0)
j <- which(x > 9)
counts <- counts[, j]

# Compute the shifted log counts.
a <- 1
s <- rowSums(counts)
s <- s / mean(s)
Y <- MatrixExtra::mapSparse(counts / (a * s), log1p)
```

## PCA and whitening

I center the transformed expression matrix and retain 25 PCs. `U` has
orthonormal columns; `X1` has whitened PCs in rows and cells in columns.

```{r pca}
n_comp <- 25
# Center within irlba to avoid creating a dense cells-by-genes matrix.
svd_fit <- irlba(Y, nv = n_comp, center = colMeans(Y))
U <- svd_fit$u
X1 <- sqrt(n) * t(U)

plot(seq_along(svd_fit$d), svd_fit$d, type = "b", pch = 16,
     xlab = "PC", ylab = "Singular value", cex.axis = 1.3, cex.lab = 1.3)
```

## Methods compared

I compare the following rank-1 methods:

- Log-Phi with $\alpha=1$ and $2$.
- FastICA with an intercept.
- First-order EB-proj with normalized Bernoulli and centered Exponential priors.
- Goldenplus EB-proj with a normalized Bernoulli prior and the ICA-type Hessian approximation.

All methods use 100 shared random starting directions.

The normalized Bernoulli prior is

$$
\mathbb{P}[Z = \sqrt{(1-p)/p}] = p, \qquad
\mathbb{P}[Z = -\sqrt{p/(1-p)}] = 1-p,
$$

and the centered Exponential prior is

$$
Z = T-1, \qquad T \sim \mathrm{Exp}(1).
$$

Both have mean zero and variance one. Bernoulli estimates $p$; the centered
Exponential prior is fixed. Goldenplus first fits the scaled Bernoulli prior
$cZ$ with fixed noise variance $c$, where $c=2/(1+\sqrt{5})$, then uses the
unit-variance prior $Z$ with free precision. Both stages use the ICA-type
Hessian approximation.

```{r method-setup, message=FALSE}
library(ggplot2)
code_files <- file.path("code", c("projection_methods_temp.R",
                                 "ebnm_normalized_bernoulli.R",
                                 "ebnm_centered_exponential.R",
                                 "ebproj_functions.R",
                                 "ebproj_newton_functions.R"))
for (file in code_files) source(file)

set.seed(1)
n_starts <- 100
maxiter <- 100
W_start <- matrix(rnorm(n_comp * n_starts), nrow = n_comp)
W_aug <- rbind(rnorm(n_starts), W_start)
W_start <- sweep(W_start, 2, sqrt(colSums(W_start^2)), "/")
W_aug <- sweep(W_aug, 2, sqrt(colSums(W_aug^2)), "/")
priors <- list(Bernoulli = ebnm_normalized_bernoulli,
               Exponential = ebnm_centered_exponential)
results <- list()
```

```{r ica}
for (alpha in c(1, 2)) {
  W <- W_start
  for (iter in 1:50) W <- fastica_update_logphi(X1, W, alpha = alpha)
  L <- t(X1) %*% W
  results[[paste("Log-Phi alpha =", alpha)]] <- list(
    loadings = L, objective = colMeans(pnorm(alpha * L, log.p = TRUE)))
}

X1_aug <- rbind(1, X1)
W <- W_aug
for (iter in 1:50) W <- fastica_update(X1_aug, W)
L <- t(X1_aug) %*% W
results[["FastICA"]] <- list(loadings = L, objective = colMeans(log(cosh(L))))
```

```{r eb-proj, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
c_golden <- 2 / (1 + sqrt(5))

# Scale posterior moments and density; fitted_g stores the unit prior's shape.
warm_prior <- function(x, s, g_init = NULL, fix_g = FALSE, output = NULL) {
  fit <- ebnm_normalized_bernoulli(x / c_golden, s / c_golden,
    g_init = g_init, fix_g = fix_g, output = output)
  fit$posterior$mean <- c_golden * fit$posterior$mean
  fit$posterior$sd <- c_golden * fit$posterior$sd
  fit$posterior$second_moment <- c_golden^2 * fit$posterior$second_moment
  fit$log_likelihood <- as.numeric(fit$log_likelihood) - length(x) * log(c_golden)
  fit
}

fit_goldplus <- function(w) {
  warm <- eb_proj_newton(U, ebnm_fn = warm_prior, w_init = w, g_init = 0.5,
                        sigma2 = c_golden, fix_tau = TRUE, maxiter = maxiter)
  eb_proj_newton(U, ebnm_fn = ebnm_normalized_bernoulli, w_init = warm$w,
                g_init = warm$fitted_g, tau = warm$tau, maxiter = maxiter)
}

L <- matrix(NA_real_, n, n_starts)
objective <- numeric(n_starts)
status <- character(n_starts)
for (j in seq_len(n_starts)) {
  fit <- fit_goldplus(W_start[, j])
  L[, j] <- fit$x
  objective[j] <- fit$objective
  status[j] <- fit$status
}
eb_results <- list("Goldplus-Bern" = list(loadings = L, objective = objective, status = status))
```

```{r first-order, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
P <- tcrossprod(U)
first_order_results <- list()
for (prior_name in names(priors)) {
  prior <- priors[[prior_name]]
  exponential <- prior_name == "Exponential"
  method <- paste0("EBproj-1st-order-", if (exponential) "Exp" else "Bern")
  L <- matrix(0, n, n_starts)
  objective <- numeric(n_starts)
  status <- character(n_starts)
  for (j in seq_len(n_starts)) {
    x_init <- as.vector(t(X1) %*% W_start[, j])
    s_init <- sqrt(n / (n - 2 * n_comp))
    initial <- if (exponential) prior(x_init, s_init) else
      prior(x_init, s_init, g_init = 0.5, fix_g = TRUE)
    fit <- eb_proj_rank1_update(P, initial$posterior$mean, ebnm_fn = prior,
      v2_init = sum(initial$posterior$second_moment),
      k = n_comp, b = n_comp / 2, maxiter = maxiter)
    direction <- as.vector(crossprod(U, fit$vbar))
    direction_norm <- sqrt(sum(direction^2))
    status[j] <- if (fit$niter < maxiter) "converged" else "maximum iterations reached"
    if (fit$v2 == 0) {
      status[j] <- "null solution"
    } else if (direction_norm <= .Machine$double.eps * max(1, sqrt(fit$v2))) {
      status[j] <- "zero projected posterior mean"
    } else {
      L[, j] <- sqrt(n) * U %*% (direction / direction_norm)
    }
    objective[j] <- fit$objective
  }
  first_order_results[[method]] <- list(loadings = L, objective = objective, status = status)
}
```

```{r collect-results}
method_order <- c("Log-Phi alpha = 1", "Log-Phi alpha = 2",
                  "Goldplus-Bern",
                  "EBproj-1st-order-Bern", "EBproj-1st-order-Exp", "FastICA")
results <- c(results, eb_results, first_order_results)[method_order]
```

## Grouped loading plots

I group solutions by complete-linkage clustering of $1-|\mathrm{cor}|$, cut at
0.1, as in [Matthew's pancreas note](https://stephenslab.github.io/single-cell-jamboree/pancreas_celseq2_ica.html).
For each cluster, the first fit is shown, and the plots are ordered by decreasing objective.
`n` counts starts; `obj` is the representative fit's objective.
Each figure holds up to 25 subplots in five columns, with fixed subplot sizes
and figure height adjusted to the occupied rows.

EB fits that stop at a null or zero projected posterior mean are excluded.

<!-- iteration-limit fits remain in the plots. -->

Cells are ordered by cell type, then treatment condition. Use the tabs under
each method to switch between cell-type and condition coloring.

First-order EB-proj panels plot $L=\sqrt{n}\,Uw$, where
$w=U^\top m/\|U^\top m\|$ and $m$ is the posterior mean before projection.

```{r plot-helper}
plot_solutions <- function(result, method, color_by, start_labels = NULL) {
  L <- result$loadings
  keep <- which(apply(L, 2, sd) > 1e-8 & is.finite(result$objective))
  if (!is.null(result$status)) {
    keep <- keep[result$status[keep] %in% c("converged", "maximum iterations reached")]
  }
  if (!length(keep)) return(invisible(NULL))
  clusters <- if (!is.null(start_labels)) seq_along(keep) else if (length(keep) == 1) 1 else cutree(
    hclust(as.dist(1 - pmin(abs(cor(L[, keep, drop = FALSE])), 1)),
           method = "complete"), h = 0.1)
  groups <- split(keep, clusters)
  representatives <- vapply(groups, function(idx) idx[1], 1L)
  o <- order(result$objective[representatives], decreasing = TRUE)
  representatives <- representatives[o]
  labels <- sprintf("n: %d  obj: %.3f", lengths(groups)[o],
                    result$objective[representatives])
  if (!is.null(start_labels)) labels <- sprintf("%s  obj: %.3f",
    start_labels[representatives], result$objective[representatives])
  cell_order <- order(sample_info$celltype, sample_info$condition)
  pages <- split(seq_along(representatives), ceiling(seq_along(representatives) / 25))
  for (page in seq_along(pages)) {
    idx <- pages[[page]]
    df <- data.frame(
      cell = rep(seq_len(n), length(idx)),
      loading = as.vector(L[cell_order, representatives[idx], drop = FALSE]),
      panel = factor(rep(representatives[idx], each = n), levels = representatives[idx]),
      celltype = rep(sample_info$celltype[cell_order], length(idx)),
      condition = rep(sample_info$condition[cell_order], length(idx)))
    p <- ggplot(df, aes(cell, loading, color = .data[[color_by]])) +
      geom_point(size = 0.5, alpha = 0.7) +
      geom_hline(yintercept = 0, linetype = 2, linewidth = 0.3) +
      facet_wrap(~ panel, ncol = 5, scales = "free_y",
                 labeller = labeller(panel = setNames(labels, representatives))) +
      scale_color_manual(values = hcl.colors(nlevels(df[[color_by]]), "Dark 3")) +
      labs(x = NULL, y = "Loading", color = color_by,
           subtitle = if (length(pages) > 1) paste("Page", page, "of", length(pages))) +
      theme_bw(base_size = 12) +
      theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
            legend.position = "bottom", legend.justification = "left",
            strip.text = element_text(size = 10, margin = margin(2, 0, 2, 0)),
            strip.background = element_rect(fill = "grey90", color = NA)) +
      guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
    g <- ggplotGrob(p)
    panels <- grepl("^panel", g$layout$name)
    g$widths[unique(g$layout$l[panels])] <- grid::unit(1.9, "in")
    g$heights[unique(g$layout$t[panels])] <- grid::unit(1.15, "in")
    g$vp <- grid::viewport(x = 0, y = 1, just = c("left", "top"),
                          width = sum(g$widths), height = sum(g$heights))
    height <- grid::convertHeight(sum(g$heights), "in", valueOnly = TRUE)
    file <- knitr::fig_path(paste0(color_by, "-", make.names(method), "-", page, ".png"))
    xfun::in_dir(knitr::opts_knit$get("output.dir"), {
      dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
      ggsave(file, g, width = 12, height = height + 0.1, units = "in", dpi = 300, bg = "white")
      cat(knitr::knit_hooks$get("plot")(file, knitr::opts_current$get()), "\n\n", sep = "")
    })
  }
}
```

```{r grouped-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "condition")) {
    heading <- if (color_by == "celltype") "Cell type" else "Condition"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(results[[method]], method, color_by)
  }
}
```

## Initialize from condition indicators

I project each of the four centered condition indicators into the retained PC
subspace and normalize its direction. First-order EB-proj-Bern and Goldplus-Bern
use these starts with the same settings as above. Each panel is labeled by its
starting condition, and panels are ordered by decreasing objective.

```{r condition-starts, cache=TRUE, cache.extra=list(U, sample_info$condition, maxiter, tools::md5sum(code_files)), dependson="eb-proj"}
L_condition <- scale(model.matrix(~ condition - 1, sample_info),
                     center = TRUE, scale = FALSE)
W_condition <- crossprod(U, L_condition)
W_condition <- sweep(W_condition, 2, sqrt(colSums(W_condition^2)), "/")
colnames(W_condition) <- levels(sample_info$condition)
condition_results <- list()
for (method in c("EBproj-1st-order-Bern", "Goldplus-Bern")) {
  L <- matrix(0, n, ncol(W_condition))
  objective <- numeric(ncol(W_condition))
  status <- character(ncol(W_condition))
  for (j in seq_len(ncol(W_condition))) {
    if (method == "Goldplus-Bern") {
      fit <- fit_goldplus(W_condition[, j])
      L[, j] <- fit$x
      status[j] <- fit$status
    } else {
      x_init <- as.vector(t(X1) %*% W_condition[, j])
      initial <- ebnm_normalized_bernoulli(x_init, sqrt(n / (n - 2 * n_comp)),
                                         g_init = 0.5, fix_g = TRUE)
      fit <- eb_proj_rank1_update(P, initial$posterior$mean,
        ebnm_fn = ebnm_normalized_bernoulli,
        v2_init = sum(initial$posterior$second_moment),
        k = n_comp, b = n_comp / 2, maxiter = maxiter)
      direction <- as.vector(crossprod(U, fit$vbar))
      direction_norm <- sqrt(sum(direction^2))
      status[j] <- if (fit$niter < maxiter) "converged" else "maximum iterations reached"
      if (fit$v2 == 0) {
        status[j] <- "null solution"
      } else if (direction_norm <= .Machine$double.eps * max(1, sqrt(fit$v2))) {
        status[j] <- "zero projected posterior mean"
      } else {
        L[, j] <- sqrt(n) * U %*% (direction / direction_norm)
      }
    }
    objective[j] <- fit$objective
  }
  condition_results[[method]] <- list(loadings = L, objective = objective, status = status)
}
```

```{r condition-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(condition_results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "condition")) {
    heading <- if (color_by == "celltype") "Cell type" else "Condition"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(condition_results[[method]], method, color_by,
                   start_labels = colnames(W_condition))
  }
}
```
