---
title: "Analysis of cytokine-stimulated mouse islet cells"
author: "Junming Guan"
date: "2026-08-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Data and preprocessing

Here I analyze the pancreas cytokine data, retaining the cell-type and treatment-condition
labels for visualization. The goal is to investigate if our proposed methods can identify cell-type or condition contrasts. 

Genes expressed in fewer than 10 cells are removed,
and counts are library-size normalized before applying the shifted log transform.

```{r setup, message=FALSE}
library(irlba)
set.seed(1)
load("data/pancreas_cytokine_lsa.RData")
barcodes <- as.data.frame(barcodes)
sample_info <- data.frame(
  celltype = factor(barcodes$celltype,
                    c("Acinar", "Ductal", "Endothelial/Mesnchymal", "Macrophage",
                      "Alpha", "Beta", "Delta", "Gamma")),
  condition = factor(barcodes$condition,
                     c("Untreated", "IL-1B", "IFNg", "IL-1B_IFNg"))
)
n <- nrow(counts)
table(sample_info$celltype, sample_info$condition)
```

```{r preprocess}
# Remove genes that are expressed in fewer than 10 cells.
x <- colSums(counts > 0)
j <- which(x > 9)
counts <- counts[, j]

# Compute the shifted log counts.
a <- 1
s <- rowSums(counts)
s <- s / mean(s)
Y <- MatrixExtra::mapSparse(counts / (a * s), log1p)
```

## PCA and whitening

I center the transformed expression matrix and retain 25 PCs. `U` has
orthonormal columns; `X1` has whitened PCs in rows and cells in columns.

```{r pca}
n_comp <- 25
# Center within irlba to avoid creating a dense cells-by-genes matrix.
svd_fit <- irlba(Y, nv = n_comp, center = colMeans(Y))
U <- svd_fit$u
X1 <- sqrt(n) * t(U)

plot(seq_along(svd_fit$d), svd_fit$d, type = "b", pch = 16,
     xlab = "PC", ylab = "Singular value", cex.axis = 1.3, cex.lab = 1.3)
```

## Methods compared

I compare the following rank-1 methods:

- Log-Phi with $\alpha=1$ and $2$.
- FastICA with an intercept.
- First-order EB-proj with normalized Bernoulli and centered Exponential priors.
- Goldenplus EB-proj with a normalized Bernoulli prior and the ICA-type Hessian approximation.

All methods use 100 shared random starting directions.

The normalized Bernoulli prior is

$$
\mathbb{P}[Z = \sqrt{(1-p)/p}] = p, \qquad
\mathbb{P}[Z = -\sqrt{p/(1-p)}] = 1-p,
$$

and the centered Exponential prior is

$$
Z = T-1, \qquad T \sim \mathrm{Exp}(1).
$$

Both have mean zero and variance one. Bernoulli estimates $p$; the centered
Exponential prior is fixed. Goldenplus first fits the scaled Bernoulli prior
$cZ$ with fixed noise variance $c$, where $c=2/(1+\sqrt{5})$, then uses the
unit-variance prior $Z$ with free precision. Both stages use the ICA-type
Hessian approximation.

```{r method-setup, message=FALSE}
library(ggplot2)
code_files <- file.path("code", c("projection_methods_temp.R",
                                 "ebnm_normalized_bernoulli.R",
                                 "ebnm_centered_exponential.R",
                                 "ebproj_functions.R",
                                 "ebproj_newton_functions.R"))
for (file in code_files) source(file)

set.seed(1)
n_starts <- 100
maxiter <- 100
W_start <- matrix(rnorm(n_comp * n_starts), nrow = n_comp)
W_aug <- rbind(rnorm(n_starts), W_start)
W_start <- sweep(W_start, 2, sqrt(colSums(W_start^2)), "/")
W_aug <- sweep(W_aug, 2, sqrt(colSums(W_aug^2)), "/")
priors <- list(Bernoulli = ebnm_normalized_bernoulli,
               Exponential = ebnm_centered_exponential)
results <- list()
```

```{r ica}
for (alpha in c(1, 2)) {
  W <- W_start
  for (iter in 1:50) W <- fastica_update_logphi(X1, W, alpha = alpha)
  L <- t(X1) %*% W
  results[[paste("Log-Phi alpha =", alpha)]] <- list(
    loadings = L, objective = colMeans(pnorm(alpha * L, log.p = TRUE)))
}

X1_aug <- rbind(1, X1)
W <- W_aug
for (iter in 1:50) W <- fastica_update(X1_aug, W)
L <- t(X1_aug) %*% W
results[["FastICA"]] <- list(loadings = L, objective = colMeans(log(cosh(L))))
```

```{r eb-proj, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
c_golden <- 2 / (1 + sqrt(5))

# Scale posterior moments and density; fitted_g stores the unit prior's shape.
warm_prior <- function(x, s, g_init = NULL, fix_g = FALSE, output = NULL) {
  fit <- ebnm_normalized_bernoulli(x / c_golden, s / c_golden,
    g_init = g_init, fix_g = fix_g, output = output)
  fit$posterior$mean <- c_golden * fit$posterior$mean
  fit$posterior$sd <- c_golden * fit$posterior$sd
  fit$posterior$second_moment <- c_golden^2 * fit$posterior$second_moment
  fit$log_likelihood <- as.numeric(fit$log_likelihood) - length(x) * log(c_golden)
  fit
}

fit_goldplus <- function(w) {
  warm <- eb_proj_newton(U, ebnm_fn = warm_prior, w_init = w, g_init = 0.5,
                        sigma2 = c_golden, fix_tau = TRUE, maxiter = maxiter)
  eb_proj_newton(U, ebnm_fn = ebnm_normalized_bernoulli, w_init = warm$w,
                g_init = warm$fitted_g, tau = warm$tau, maxiter = maxiter)
}

L <- matrix(NA_real_, n, n_starts)
objective <- numeric(n_starts)
status <- character(n_starts)
for (j in seq_len(n_starts)) {
  fit <- fit_goldplus(W_start[, j])
  L[, j] <- fit$x
  objective[j] <- fit$objective
  status[j] <- fit$status
}
eb_results <- list("Goldplus-Bern" = list(loadings = L, objective = objective, status = status))
```

```{r first-order, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
P <- tcrossprod(U)
first_order_results <- list()
for (prior_name in names(priors)) {
  prior <- priors[[prior_name]]
  exponential <- prior_name == "Exponential"
  method <- paste0("EBproj-1st-order-", if (exponential) "Exp" else "Bern")
  L <- matrix(0, n, n_starts)
  objective <- numeric(n_starts)
  status <- character(n_starts)
  for (j in seq_len(n_starts)) {
    x_init <- as.vector(t(X1) %*% W_start[, j])
    s_init <- sqrt(n / (n - 2 * n_comp))
    initial <- if (exponential) prior(x_init, s_init) else
      prior(x_init, s_init, g_init = 0.5, fix_g = TRUE)
    fit <- eb_proj_rank1_update(P, initial$posterior$mean, ebnm_fn = prior,
      v2_init = sum(initial$posterior$second_moment),
      k = n_comp, b = n_comp / 2, maxiter = maxiter)
    direction <- as.vector(crossprod(U, fit$vbar))
    direction_norm <- sqrt(sum(direction^2))
    status[j] <- if (fit$niter < maxiter) "converged" else "maximum iterations reached"
    if (fit$v2 == 0) {
      status[j] <- "null solution"
    } else if (direction_norm <= .Machine$double.eps * max(1, sqrt(fit$v2))) {
      status[j] <- "zero projected posterior mean"
    } else {
      L[, j] <- sqrt(n) * U %*% (direction / direction_norm)
    }
    objective[j] <- fit$objective
  }
  first_order_results[[method]] <- list(loadings = L, objective = objective, status = status)
}
```

```{r collect-results}
method_order <- c("Log-Phi alpha = 1", "Log-Phi alpha = 2",
                  "Goldplus-Bern",
                  "EBproj-1st-order-Bern", "EBproj-1st-order-Exp", "FastICA")
results <- c(results, eb_results, first_order_results)[method_order]
```

## Grouped loading plots

I group solutions by complete-linkage clustering of $1-|\mathrm{cor}|$, cut at
0.1, as in [Matthew's pancreas note](https://stephenslab.github.io/single-cell-jamboree/pancreas_celseq2_ica.html).
For each cluster, the first fit is shown, and the plots are ordered by decreasing objective.
`n` counts starts; `obj` is the representative fit's objective.
Each figure holds up to 15 subplots in three columns, with fixed subplot sizes.

EB fits that stop at a null or zero projected posterior mean are excluded.

<!-- iteration-limit fits remain in the plots. -->

Cells are ordered by cell type, then treatment condition. Each set of components
is colored by cell type and by condition, keeping the same cell order in both plots.

First-order EB-proj panels plot $L=\sqrt{n}\,Uw$, where
$w=U^\top m/\|U^\top m\|$ and $m$ is the posterior mean before projection.

```{r plot-helper}
plot_solutions <- function(result, method, start_labels = NULL) {
  L <- result$loadings
  keep <- which(apply(L, 2, sd) > 1e-8 & is.finite(result$objective))
  if (!is.null(result$status)) {
    keep <- keep[result$status[keep] %in% c("converged", "maximum iterations reached")]
  }
  if (!length(keep)) return(invisible(NULL))
  clusters <- if (!is.null(start_labels)) seq_along(keep) else if (length(keep) == 1) 1 else cutree(
    hclust(as.dist(1 - pmin(abs(cor(L[, keep, drop = FALSE])), 1)),
           method = "complete"), h = 0.1)
  groups <- split(keep, clusters)
  representatives <- vapply(groups, function(idx) idx[1], 1L)
  o <- order(result$objective[representatives], decreasing = TRUE)
  representatives <- representatives[o]
  labels <- sprintf("n: %d  obj: %.3f", lengths(groups)[o],
                    result$objective[representatives])
  if (!is.null(start_labels)) labels <- sprintf("%s  obj: %.3f",
    start_labels[representatives], result$objective[representatives])
  cell_order <- order(sample_info$celltype, sample_info$condition)
  pages <- split(seq_along(representatives), ceiling(seq_along(representatives) / 15))
  for (page in seq_along(pages)) {
    idx <- pages[[page]]
    df <- data.frame(
      cell = rep(seq_len(n), length(idx)),
      loading = as.vector(L[cell_order, representatives[idx], drop = FALSE]),
      panel = factor(rep(representatives[idx], each = n), levels = representatives[idx]),
      celltype = rep(sample_info$celltype[cell_order], length(idx)),
      condition = rep(sample_info$condition[cell_order], length(idx)))
    for (color_by in c("celltype", "condition")) {
      p <- ggplot(df, aes(cell, loading, color = .data[[color_by]])) +
        geom_point(size = 0.5, alpha = 0.7) +
        geom_hline(yintercept = 0, linetype = 2, linewidth = 0.3) +
        facet_wrap(~ panel, ncol = 3, scales = "free_y",
                   labeller = labeller(panel = setNames(labels, representatives))) +
        scale_color_manual(values = hcl.colors(nlevels(df[[color_by]]), "Dark 3")) +
        labs(x = NULL, y = "Loading", color = color_by,
             title = paste(method, "-", color_by),
             subtitle = paste("Page", page, "of", length(pages))) +
        theme_bw(base_size = 16) +
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              legend.position = "bottom", strip.text = element_text(size = 14)) +
        guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
      grid::grid.newpage()
      g <- ggplotGrob(p)
      panels <- grepl("^panel", g$layout$name)
      g$widths[unique(g$layout$l[panels])] <- grid::unit(3.1, "in")
      g$heights[unique(g$layout$t[panels])] <- grid::unit(1.15, "in")
      g$vp <- grid::viewport(x = 0, y = 1, just = c("left", "top"),
                            width = sum(g$widths), height = sum(g$heights))
      grid::grid.draw(g)
    }
  }
}
```

```{r grouped-solutions, fig.width=12, fig.height=10.5}
for (method in names(results)) plot_solutions(results[[method]], method)
```
