---
title: "Shifted Rademacher prior for EBproj"
author: "junmingguan"
date: "2026-06-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Matthew's results suggest that if the data matrix with binary sources is centered, then it's no longer appropriate to put a Rademacher type prior. Instead we should explicitly model the shift due to centering. Consider the model:

$$
x_i \mid z_i, s_i^2 \sim N(z_i, s_i^2),
\qquad
z_i \sim \text{GSRad}(\pi, c):=(1-\pi)\delta_{-1+c} + \pi\delta_{1+c},
$$
where GSRad stands for "generalized shifted Rademacher". The prior proportion $\pi$ and the shift $c$ shall be estimated from the data by maximizing the marginal log likelihood:

$$
\ell(c,\pi)
=
\sum_{i=1}^n
\log
\left[
(1-\pi)\,\phi(x_i; c-1, s_i^2)
+
\pi\,\phi(x_i; c+1, s_i^2)
\right].
$$
To facilitate optimization, consider the reparametrization of $\pi$, so that we don't have to worry about boundary cases:

$$
\eta = \operatorname{logit}(\pi),
\qquad
\pi = \operatorname{logit}^{-1}(\eta)
=
\frac{1}{1+\exp(-\eta)}.
$$
Then the marginal log likelihood is given by

$$
\ell(c,\eta)
=
\sum_{i=1}^n
\log
\left[
\frac{\exp(-\eta)}{1+\exp(-\eta)}
\phi(x_i; c-1, s_i^2)
+
\frac{1}{1+\exp(-\eta)}
\phi(x_i; c+1, s_i^2)
\right].
$$
If $\pi$ is fixed at $1/2$ and $c$ is to be estimated, the prior is termed shifted Rademacher: $\text{SRad}(c)$.

For a centered and whitened asymmetric binary source, it is natural to use a two-point prior whose mean and variance are fixed to match the transformed source. We call this the normalized Rademacher prior:

$$
z_i \sim \text{NRad}(\pi, c)
=
\pi \delta_{c - a_\pi}
+
(1-\pi)\delta_{c + b_\pi},
\qquad
a_\pi = \sqrt{\frac{1-\pi}{\pi}},
\qquad
b_\pi = \sqrt{\frac{\pi}{1-\pi}}.
$$

This prior satisfies

$$
\mathbb E[z_i] = c,
\qquad
\operatorname{Var}(z_i)=1.
$$

Its marginal log likelihood is

$$
\ell(c,\pi)
=
\sum_{i=1}^n
\log
\left[
\pi\,\phi(x_i; c-a_\pi, s_i^2)
+
(1-\pi)\,\phi(x_i; c+b_\pi, s_i^2)
\right].
$$

The posterior is again a two-point distribution. Let

$$
\theta_- = c-a_\pi,
\qquad
\theta_+ = c+b_\pi,
\qquad
r_i = \Pr(z_i=\theta_+ \mid x_i).
$$

Then

$$
r_i
=
\frac{
(1-\pi)\phi(x_i; \theta_+, s_i^2)
}{
\pi\phi(x_i; \theta_-, s_i^2)
+
(1-\pi)\phi(x_i; \theta_+, s_i^2)
},
$$

and the posterior moments are

$$
\mathbb E[z_i \mid x_i]
=
(1-r_i)\theta_- + r_i\theta_+,
\qquad
\mathbb E[z_i^2 \mid x_i]
=
(1-r_i)\theta_-^2 + r_i\theta_+^2.
$$

Equivalently, if

$$
m_\pi = \frac{\theta_-+\theta_+}{2},
\qquad
h_\pi = \frac{\theta_+-\theta_-}{2}
=
\frac{1}{2\sqrt{\pi(1-\pi)}},
$$

then

$$
\mathbb E[z_i \mid x_i]
=
m_\pi
+
h_\pi
\tanh\left\{
\frac12\log\frac{1-\pi}{\pi}
+
\frac{x_i-m_\pi}{2s_i^2\sqrt{\pi(1-\pi)}}
\right\}.
$$


## First look at shifted (generalized) Rademacher


```{r}
source(file.path("code", "ebnm_shifted_rademacher.R"))
source(file.path("code", "ebnm_normalized_rademacher.R"))
source(file.path("code", "ebnm_rademacher.R"))
source(file.path("code", "ica_functions.R"))

ebnm_shifted_rademacher_pi02 <- ebnm_shifted_rademacher_fixed_pi(0.2)
ebnm_normalized_rademacher_c0 <- ebnm_normalized_rademacher_fixed_c(0)
```

We look at a 9-group unbalanced example.

```{r}
K=9
p = 1000
n = 100
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,20),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

X = L %*% t(FF) + rnorm(n*p,0,1)
```


We center and whiten the data.

```{r}
Xcenter = scale(X,scale=FALSE)
Ucenter = svd(Xcenter)$u[,1:K] 
Pcenter = Ucenter %*% t(Ucenter)
# equivalent to the following
X1 = preprocess(X, n.comp = K)
# P = t(X1) %*% X1 / n
```

Some note on the whitening step:

$$
\begin{aligned}
X_c &= X(I_n - \boldsymbol{11}^\top/n) = UDV^\top \\
\widetilde{X} &= X_c K = X_c V_K D_K^{-1} = U_K.
\end{aligned}
$$

```{r}
ebproj_rank1_update <- function(P, v_init, ebnm_fn = ebnm_shifted_rademacher,
                                k = NULL, b = 0, maxiter = 100,
                                tol = 1e-8, verbose = FALSE) {
  n <- nrow(P)
  if (ncol(P) != n) {
    stop("P must be a square matrix.")
  }
  if (length(v_init) != n) {
    stop("v_init must have length nrow(P).")
  }
  if (is.null(k)) {
    k <- round(sum(diag(P)))
  }
  if (n - k - 2 * b <= 0) {
    stop("n - k - 2 * b must be positive.")
  }

  vbar <- as.vector(v_init)
  v2 <- sum(vbar^2)
  tau <- Inf
  rho_trace <- rep(NA_real_, maxiter)
  tau_trace <- rep(NA_real_, maxiter)
  c_trace <- rep(NA_real_, maxiter)

  for (iter in seq_len(maxiter)) {
    Pv <- as.vector(P %*% vbar)
    rho <- as.numeric(crossprod(vbar, Pv) / v2)
    rho <- min(max(rho, 1e-12), 1 - 1e-12)

    tau <- 1 / (1 - rho)
    sigma_eff2 <- n / ((n - k - 2 * b) * tau)

    x <- Pv / rho
    s <- sqrt(sigma_eff2 * v2 / (n * rho))
    fit <- ebnm_fn(x, s = s)

    vbar_new <- fit$posterior$mean
    v2_new <- sum(fit$posterior$second_moment)

    rho_new <- as.numeric(crossprod(vbar_new, P %*% vbar_new) / v2_new)
    rho_new <- min(max(rho_new, 1e-12), 1 - 1e-12)
    rho_trace[iter] <- rho_new
    tau_trace[iter] <- tau
    c_trace[iter] <- fit$fitted_g$c

    if (verbose) {
      message(sprintf(
        "iter %d: rho = %.6f, tau = %.6f, c = %.6f",
        iter, rho_new, tau, fit$fitted_g$c
      ))
    }

    if (sqrt(sum((vbar_new - vbar)^2)) <= tol * (sqrt(sum(vbar^2)) + 1)) {
      vbar <- vbar_new
      v2 <- v2_new
      break
    }

    vbar <- vbar_new
    v2 <- v2_new
  }

  list(
    posterior = fit$posterior,
    fitted_g = fit$fitted_g,
    log_likelihood = fit$log_likelihood,
    vbar = vbar,
    v2 = v2,
    rho = as.numeric(crossprod(vbar, P %*% vbar) / v2),
    tau = 1 / (1 - as.numeric(crossprod(vbar, P %*% vbar) / v2)),
    rho_trace = rho_trace[seq_len(iter)],
    tau_trace = tau_trace[seq_len(iter)],
    c_trace = c_trace[seq_len(iter)],
    niter = iter
  )
}
```

EBproj with shifted generalized Radmacher perfectly recovers a true source, with very high alignment. At the same time, the prior parameters are also almost perfectly estimated.

```{r}
set.seed(2)
v.init = sample(c(-1,1),n,replace=TRUE)
maxiter = 100
fit.shifted = ebproj_rank1_update(
  P = Pcenter,
  v_init = v.init,
  ebnm_fn = ebnm_generalized_shifted_rademacher,
  k = 1,
  maxiter = maxiter
)
fit.shifted$rho
fit.shifted$fitted_g
cor(fit.shifted$vbar, L)
plot(fit.shifted$vbar)
```


I also tried $\text{GSRad}(0.2, c)$ where $\pi=0.2$ is fixed. The results are similar.

```{r}
fit.shifted.pi02 = ebproj_rank1_update(
  P = Pcenter,
  v_init = v.init,
  ebnm_fn = ebnm_shifted_rademacher_pi02,
  k = 1,
  maxiter = maxiter
)
fit.shifted.pi02$rho
fit.shifted.pi02$fitted_g
cor(fit.shifted.pi02$vbar, L)
plot(fit.shifted.pi02$vbar)
```

Now try the normalized Rademacher prior $\text{NRad}(\pi,c)$, where the atom locations change with $\pi$ to keep prior variance equal to one.

```{r}
fit.normalized = ebproj_rank1_update(
  P = Pcenter,
  v_init = v.init,
  ebnm_fn = ebnm_normalized_rademacher,
  k = 1,
  maxiter = maxiter
)
fit.normalized$rho
fit.normalized$fitted_g
cor(fit.normalized$vbar, L)
plot(fit.normalized$vbar)
```

If the centering step is expected to remove the source mean, we can fix the normalized prior shift at $c=0$ and estimate only $\pi$.

```{r}
fit.normalized.c0 = ebproj_rank1_update(
  P = Pcenter,
  v_init = v.init,
  ebnm_fn = ebnm_normalized_rademacher_c0,
  k = 1,
  maxiter = maxiter
)
fit.normalized.c0$rho
fit.normalized.c0$fitted_g
cor(fit.normalized.c0$vbar, L)
plot(fit.normalized.c0$vbar)
```


It would also be interesting to see what happens if we apply "EB ICA" with the shifted Rademacher. It appears that the solution is close to binary but with a bit of "noise". Also the fitted prior proportion and shift are not as accurate as in EBproj. Is it because the low-rank **plus iid noise** model is misspecified?

```{r}
fit.ebica = ebica_generalized_parallel(X1, K = 1, ebnm_fn = ebnm_generalized_shifted_rademacher, S_init = v.init, s= 1)
cor(fit.ebica$S_plus, L)
plot(fit.ebica$S_plus)
fit.ebica$fitted_priors
```
