---
title: "Simulations: cell types with donor effects"
author: "Junming Guan"
date: "2026-08-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Simulation

Here I aim to look at a simulation with similar strcuture to the pancreas dataset. I simulate the expressions for eight cell types and four donors, with gaussian noise:

$$
X = L_{\mathrm{type}} F_{\mathrm{type}}^\top
  + L_{\mathrm{donor}} F_{\mathrm{donor}}^\top + E.
$$

Every cell type contains all four donors in equal proportions. Donor effects
are shared across cell types. Both loading matrices contain centered indicators.

```{r setup}
set.seed(1)
p <- 1000
donor_sd <- 0.25
noise_sd <- 1

# Counts per donor; the last group stands in for rare cell types.
type_counts <- c(alpha = 222, beta = 120, acinar = 72, ductal = 66,
                 delta = 54, gamma = 30, stellate = 24, rare = 12)
sample_info <- data.frame(
  celltype = factor(rep(rep(names(type_counts), type_counts), 4),
                    levels = names(type_counts)),
  donor = factor(rep(paste0("D", 1:4), each = sum(type_counts)))
)
n <- nrow(sample_info)
table(sample_info$celltype, sample_info$donor)
```

I scale the data to give cell-type variance 1 and donor variance 0.25.

```{r generate-data}
L_type <- scale(model.matrix(~ celltype - 1, sample_info), scale = FALSE)
L_donor <- scale(model.matrix(~ donor - 1, sample_info), scale = FALSE)

F_type <- matrix(rnorm(p * ncol(L_type)), nrow = p)
F_donor <- matrix(rnorm(p * ncol(L_donor)), nrow = p)
F_type <- F_type / sqrt(mean((L_type %*% t(F_type))^2))
F_donor <- donor_sd * F_donor / sqrt(mean((L_donor %*% t(F_donor))^2))

X_type <- L_type %*% t(F_type)
X_donor <- L_donor %*% t(F_donor)
E <- matrix(rnorm(n * p, sd = noise_sd), nrow = n)
X <- X_type + X_donor + E
```

## Check the signal strengths

Donor signal is 6.25% of cell-type signal.

```{r signal-strengths}
data.frame(
  component = c("Cell type", "Donor", "Noise"),
  rms = c(sqrt(mean(X_type^2)), sqrt(mean(X_donor^2)), sqrt(mean(E^2)))
)
sum(X_donor^2) / sum(X_type^2)
```

## PCA and whitening

I retain 25 PCs whereas the the signal has at most 10 dimensions.
<!-- `U` has orthonormal columns; `X1` has PCs in rows and cells -->
<!-- in columns. The signal has at most ten dimensions. -->

```{r pca}
X_centered <- scale(X, scale = FALSE)
n_comp <- 25
svd_fit <- svd(X_centered, nu = n_comp, nv = 0)
U <- svd_fit$u
X1 <- sqrt(n) * t(U)

plot(seq_len(40), svd_fit$d[1:40], type = "b", pch = 16,
     xlab = "PC", ylab = "Singular value", cex.axis = 1.3, cex.lab = 1.3)
abline(v = 10, lty = 2, col = "grey50")
```

## Rank-1 experiments

I compare the following rank-1 methods:
- Log-Phi with $\alpha=1$ and $2$
- ICA with contrasts $x^3$ and $x|x|/2$
- ICA with integrated log-cosh, as Matthew suggested
- FastICA with an intercept
- First-order EB-proj with normalized Bernoulli, and centered Exponential.
- Goldenplus EB-proj with normalized Bernoulli and normalized point-Exponential,
  both using the ICA-type Hessian approximation.

I look at 100 random starts and visualize their solutions.

Integrated log-cosh uses $g(x)=\log\cosh(x)$ and $g'(x)=\tanh(x)$ to smooth
the $x|x|/2$ contrast. For objectives and GS ordering, I approximate
$G(x)=\int_0^x\log\cosh(t)\,dt$ on a trapezoidal grid with spacing 0.001.

The normalized Bernoulli is given by

$$
\mathbb{P}[z = \sqrt{(1-p)/p}] = p \quad \mathbb{P}[z = -\sqrt{p/(1-p)}] = 1-p,
$$

where centered Exponential is defined as

$$
Z = T-1,\quad T\sim \text{Exp}(1).
$$

Both have mean $0$ and variance $1$. Note that EB-proj with centered Exp is not really an EB method, as the prior is fixed.

For point-Exponential, $Z_q=(BT-q)/\sqrt{q(2-q)}$, where
$B\sim\mathrm{Bernoulli}(q)$ and $T\sim\mathrm{Exp}(1)$ are independent.
I estimate $q$, the slab probability, starting at 0.5. Goldenplus uses prior
$cZ_q$ and fixed noise variance $c=2/(1+\sqrt{5})$ for warm-up, then prior
$Z_q$ and free precision, with at most 100 Newton updates per phase.

<!-- Use 100 shared random starts, 50 Log-Phi/FastICA updates, and at most 100
updates per EB stage. FastICA uses an intercept; Log-Phi and EB-proj use
centered projections. Bernoulli estimates $\pi$ in
$Z=(B-\pi)/\sqrt{\pi(1-\pi)}$. Exponential fixes $Z=T-1$, $T\sim\mathrm{Exp}(1)$.

For Goldplus-Bern, warm up with prior $cZ$ and fixed noise variance $c$, where
$c=2/(1+\sqrt{5})$, then run Newton with the unit-variance prior $Z$ and free
precision. Both stages use `hessian = "ica"`.
The first-order fits use our `eb_proj_rank1_update()`, initialized with
posterior moments at $\tau=1$ along the same starting directions. -->

```{r method-setup, message=FALSE}
library(ggplot2)
code_files <- file.path("code", c("projection_methods_temp.R",
                                 "ebnm_normalized_bernoulli.R",
                                 "ebnm_centered_exponential.R",
                                 "ebnm_normalized_point_exponential.R",
                                 "ebproj_functions.R",
                                 "ebproj_newton_functions.R"))
for (file in code_files) source(file)

set.seed(1)
n_starts <- 100
maxiter <- 100
W_start <- matrix(rnorm(n_comp * n_starts), nrow = n_comp)
W_aug <- rbind(rnorm(n_starts), W_start)
W_start <- sweep(W_start, 2, sqrt(colSums(W_start^2)), "/")
W_aug <- sweep(W_aug, 2, sqrt(colSums(W_aug^2)), "/")
priors <- list(Bernoulli = ebnm_normalized_bernoulli,
               Exponential = ebnm_centered_exponential)
results <- list()
```

```{r ica}
for (alpha in c(1, 2)) {
  W <- W_start
  for (iter in 1:50) W <- fastica_update_logphi(X1, W, alpha = alpha)
  L <- t(X1) %*% W
  results[[paste("Log-Phi alpha =", alpha)]] <- list(
    loadings = L, objective = colMeans(pnorm(alpha * L, log.p = TRUE)))
}

W <- W_start
for (iter in 1:50) W <- fastica_update_cubic(X1, W)
L <- t(X1) %*% W
results[["ICA x^3"]] <- list(loadings = L, objective = colMeans(L^3))

W <- W_start
for (iter in 1:50) W <- fastica_update_skew(X1, W)
L <- t(X1) %*% W
results[["ICA x|x|/2"]] <- list(loadings = L, objective = colMeans(L * abs(L) / 2))

W <- W_start
for (iter in 1:50) W <- fastica_update_ilogcosh(X1, W)
L <- t(X1) %*% W
results[["ICA integrated log-cosh"]] <- list(
  loadings = L, objective = colMeans(integrated_logcosh(L)))

X1_aug <- rbind(1, X1)
W <- W_aug
for (iter in 1:50) W <- fastica_update(X1_aug, W)
L <- t(X1_aug) %*% W
results[["FastICA"]] <- list(loadings = L, objective = colMeans(log(cosh(L))))
```

```{r eb-proj, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
c_golden <- 2 / (1 + sqrt(5))

# Scale posterior moments and density; fitted_g stores the unit prior's shape.
warm_prior <- function(x, s, g_init = NULL, fix_g = FALSE, output = NULL) {
  fit <- ebnm_normalized_bernoulli(x / c_golden, s / c_golden,
    g_init = g_init, fix_g = fix_g, output = output)
  fit$posterior$mean <- c_golden * fit$posterior$mean
  fit$posterior$sd <- c_golden * fit$posterior$sd
  fit$posterior$second_moment <- c_golden^2 * fit$posterior$second_moment
  fit$log_likelihood <- as.numeric(fit$log_likelihood) - length(x) * log(c_golden)
  fit
}

fit_goldplus <- function(w) {
  warm <- eb_proj_newton(U, ebnm_fn = warm_prior, w_init = w, g_init = 0.5,
                        sigma2 = c_golden, fix_tau = TRUE, maxiter = maxiter)
  eb_proj_newton(U, ebnm_fn = ebnm_normalized_bernoulli, w_init = warm$w,
                g_init = warm$fitted_g, tau = warm$tau, maxiter = maxiter)
}

L <- matrix(NA_real_, n, n_starts)
objective <- numeric(n_starts)
status <- character(n_starts)
for (j in seq_len(n_starts)) {
  fit <- fit_goldplus(W_start[, j])
  L[, j] <- fit$x
  objective[j] <- fit$objective
  status[j] <- fit$status
}
eb_results <- list("Goldplus-Bern" = list(loadings = L, objective = objective, status = status))
```

```{r point-exp-goldenplus, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
L <- matrix(NA_real_, n, n_starts)
objective <- q <- numeric(n_starts)
status <- character(n_starts)
for (j in seq_len(n_starts)) {
  fit <- eb_proj_goldenplus_point_exponential(U, w_init = W_start[, j], maxiter = maxiter)
  L[, j] <- fit$x
  objective[j] <- fit$objective
  q[j] <- fit$fitted_g$q
  status[j] <- fit$status
}
point_exp_results <- list("Goldplus-PointExp" = list(
  loadings = L, objective = objective, status = status, q = q))
```

```{r first-order, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
P <- tcrossprod(U)
first_order_results <- list()
for (prior_name in names(priors)) {
  prior <- priors[[prior_name]]
  exponential <- prior_name == "Exponential"
  method <- paste0("EBproj-1st-order-", if (exponential) "Exp" else "Bern")
  L <- matrix(0, n, n_starts)
  objective <- numeric(n_starts)
  status <- character(n_starts)
  for (j in seq_len(n_starts)) {
    x_init <- as.vector(t(X1) %*% W_start[, j])
    s_init <- sqrt(n / (n - 2 * n_comp))
    initial <- if (exponential) prior(x_init, s_init) else
      prior(x_init, s_init, g_init = 0.5, fix_g = TRUE)
    fit <- eb_proj_rank1_update(P, initial$posterior$mean, ebnm_fn = prior,
      v2_init = sum(initial$posterior$second_moment),
      k = n_comp, b = n_comp / 2, maxiter = maxiter)
    direction <- as.vector(crossprod(U, fit$vbar))
    direction_norm <- sqrt(sum(direction^2))
    status[j] <- if (fit$niter < maxiter) "converged" else "maximum iterations reached"
    if (fit$v2 == 0) {
      status[j] <- "null solution"
    } else if (direction_norm <= .Machine$double.eps * max(1, sqrt(fit$v2))) {
      status[j] <- "zero projected posterior mean"
    } else {
      L[, j] <- sqrt(n) * U %*% (direction / direction_norm)
    }
    objective[j] <- fit$objective
  }
  first_order_results[[method]] <- list(loadings = L, objective = objective, status = status)
}
```

```{r collect-results}
method_order <- c("Log-Phi alpha = 1", "Log-Phi alpha = 2",
                  "ICA x^3", "ICA x|x|/2", "ICA integrated log-cosh",
                  "Goldplus-Bern", "Goldplus-PointExp",
                  "EBproj-1st-order-Bern", "EBproj-1st-order-Exp", "FastICA")
results <- c(results, eb_results, point_exp_results, first_order_results)[method_order]
```

## Grouped loading plots

Here I group solutions using the method in [Matthew's pancreas note](https://stephenslab.github.io/single-cell-jamboree/pancreas_celseq2_ica.html). For each cluster, the first fit is shown, and the plots are ordered by decreasing objective.
`n` counts starts; `obj` is the representative fit's objective.
Each figure holds up to 25 subplots in five columns, as in Matthew's note,
with fixed subplot sizes and figure height adjusted to the occupied rows.

<!-- EB fits that stop at a null or zero projected -->
<!-- posterior mean are excluded; iteration-limit fits remain in the plots. -->

Cells are ordered by cell type, then donor. Use the tabs under each method
to switch between cell-type and donor coloring.

First-order EB-proj panels plot $L=\sqrt{n}\,Uw$, where
$w=U^\top m/\|U^\top m\|$ and $m$ is the posterior mean before projection.

```{r plot-helper}
plot_solutions <- function(result, method, color_by, start_labels = NULL) {
  L <- result$loadings
  keep <- if (!is.null(start_labels)) seq_len(ncol(L)) else
    which(apply(L, 2, sd) > 1e-8 & is.finite(result$objective))
  if (!is.null(result$status)) {
    keep <- keep[result$status[keep] %in% c("converged", "maximum iterations reached")]
  }
  if (!length(keep)) return(invisible(NULL))
  clusters <- if (!is.null(start_labels)) seq_along(keep) else if (length(keep) == 1) 1 else cutree(
    hclust(as.dist(1 - pmin(abs(cor(L[, keep, drop = FALSE])), 1)),
           method = "complete"), h = 0.1)
  groups <- split(keep, clusters)
  representatives <- vapply(groups, function(idx) idx[1], 1L)
  o <- if (is.null(result$objective)) seq_along(representatives) else
    order(result$objective[representatives], decreasing = TRUE)
  representatives <- representatives[o]
  if (is.null(result$objective)) {
    labels <- start_labels[representatives]
  } else {
    labels <- sprintf("n: %d  obj: %.3f", lengths(groups)[o],
                      result$objective[representatives])
    if (!is.null(start_labels)) labels <- sprintf("%s  obj: %.3f",
      start_labels[representatives], result$objective[representatives])
  }
  cell_order <- order(sample_info$celltype, sample_info$donor)
  pages <- split(seq_along(representatives), ceiling(seq_along(representatives) / 25))
  for (page in seq_along(pages)) {
    idx <- pages[[page]]
    df <- data.frame(
      cell = rep(seq_len(n), length(idx)),
      loading = as.vector(L[cell_order, representatives[idx], drop = FALSE]),
      panel = factor(rep(representatives[idx], each = n), levels = representatives[idx]),
      celltype = rep(sample_info$celltype[cell_order], length(idx)),
      donor = rep(sample_info$donor[cell_order], length(idx)))
    p <- ggplot(df, aes(cell, loading, color = .data[[color_by]])) +
      geom_point(size = 0.5, alpha = 0.7) +
      geom_hline(yintercept = 0, linetype = 2, linewidth = 0.3) +
      facet_wrap(~ panel, ncol = 5, scales = "free_y",
                 labeller = labeller(panel = setNames(labels, representatives))) +
      scale_color_manual(values = hcl.colors(nlevels(df[[color_by]]), "Dark 3")) +
      labs(x = NULL, y = "Loading", color = color_by,
           subtitle = if (length(pages) > 1) paste("Page", page, "of", length(pages))) +
      theme_bw(base_size = 12) +
      theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
            legend.position = "bottom", legend.justification = "left",
            strip.text = element_text(size = 10, margin = margin(2, 0, 2, 0)),
            strip.background = element_rect(fill = "grey90", color = NA)) +
      guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
    g <- ggplotGrob(p)
    panels <- grepl("^panel", g$layout$name)
    g$widths[unique(g$layout$l[panels])] <- grid::unit(1.9, "in")
    g$heights[unique(g$layout$t[panels])] <- grid::unit(1.15, "in")
    g$vp <- grid::viewport(x = 0, y = 1, just = c("left", "top"),
                          width = sum(g$widths), height = sum(g$heights))
    height <- grid::convertHeight(sum(g$heights), "in", valueOnly = TRUE)
    file <- knitr::fig_path(paste0(color_by, "-", make.names(method), "-", page, ".png"))
    xfun::in_dir(knitr::opts_knit$get("output.dir"), {
      dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
      ggsave(file, g, width = 12, height = height + 0.1, units = "in", dpi = 300, bg = "white")
      cat(knitr::knit_hooks$get("plot")(file, knitr::opts_current$get()), "\n\n", sep = "")
    })
  }
}
```

```{r grouped-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "donor")) {
    heading <- if (color_by == "celltype") "Cell type" else "Donor"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(results[[method]], method, color_by)
  }
}
```


## Initialize from the true donor loadings

Here I try initialize EB-proj-NormalizedBernoulli (both first order and goldenplus Newton) from the four donor loadings.

<!-- Project each of the four centered donor indicators in `L_donor` into the
retained PC subspace and normalize its direction. Fit both Bernoulli methods
from these four starts, using the same settings as above. Keep one panel per
starting donor, ordered by decreasing objective. -->

```{r donor-starts, cache=TRUE, cache.extra=list(U, L_donor, maxiter, tools::md5sum(code_files)), dependson="eb-proj"}
W_donor <- crossprod(U, L_donor)
W_donor <- sweep(W_donor, 2, sqrt(colSums(W_donor^2)), "/")
colnames(W_donor) <- levels(sample_info$donor)
donor_results <- list()
for (method in c("EBproj-1st-order-Bern", "Goldplus-Bern")) {
  L <- matrix(0, n, ncol(W_donor))
  objective <- numeric(ncol(W_donor))
  status <- character(ncol(W_donor))
  for (j in seq_len(ncol(W_donor))) {
    if (method == "Goldplus-Bern") {
      fit <- fit_goldplus(W_donor[, j])
      L[, j] <- fit$x
      status[j] <- fit$status
    } else {
      x_init <- as.vector(t(X1) %*% W_donor[, j])
      initial <- ebnm_normalized_bernoulli(x_init, sqrt(n / (n - 2 * n_comp)),
                                         g_init = 0.5, fix_g = TRUE)
      fit <- eb_proj_rank1_update(P, initial$posterior$mean,
        ebnm_fn = ebnm_normalized_bernoulli,
        v2_init = sum(initial$posterior$second_moment),
        k = n_comp, b = n_comp / 2, maxiter = maxiter)
      direction <- as.vector(crossprod(U, fit$vbar))
      direction_norm <- sqrt(sum(direction^2))
      status[j] <- if (fit$niter < maxiter) "converged" else "maximum iterations reached"
      if (fit$v2 == 0) {
        status[j] <- "null solution"
      } else if (direction_norm <= .Machine$double.eps * max(1, sqrt(fit$v2))) {
        status[j] <- "zero projected posterior mean"
      } else {
        L[, j] <- sqrt(n) * U %*% (direction / direction_norm)
      }
    }
    objective[j] <- fit$objective
  }
  donor_results[[method]] <- list(loadings = L, objective = objective, status = status)
}
```

```{r donor-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(donor_results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "donor")) {
    heading <- if (color_by == "celltype") "Cell type" else "Donor"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(donor_results[[method]], method, color_by,
                   start_labels = colnames(W_donor))
  }
}
```



## Rank-r (25) expriments

### rank-r ICA

I fit rank-25 ICA with Log-Phi ($\alpha=1,2$), $x^3$, $x|x|/2$, and integrated
log-cosh, each with parallel (polar) orthogonalization and Gram-Schmidt (GS).

 <!-- All ten fits use the same
orthonormal start and 500 updates. Each figure shows all 25 components, without
clustering, ordered by decreasing objective. -->

```{r rankr-ica, cache=TRUE, cache.extra=list(X1, tools::md5sum(code_files))}
r <- 25
n_iter <- 500
set.seed(2)
W_rankr <- polar(matrix(rnorm(n_comp * r), n_comp, r))
rankr_methods <- list(
  "Log-Phi alpha = 1" = list(
    parallel = function(U, W) fastica_update_logphi_rankr(U, W, alpha = 1),
    GS = function(U, W) fastica_update_logphi_rankr_gs(U, W, alpha = 1),
    objective = function(L) colMeans(pnorm(L, log.p = TRUE))),
  "Log-Phi alpha = 2" = list(
    parallel = function(U, W) fastica_update_logphi_rankr(U, W, alpha = 2),
    GS = function(U, W) fastica_update_logphi_rankr_gs(U, W, alpha = 2),
    objective = function(L) colMeans(pnorm(2 * L, log.p = TRUE))),
  "ICA x^3" = list(
    parallel = fastica_update_cubic_rankr,
    GS = fastica_update_cubic_rankr_gs,
    objective = function(L) colMeans(L^3)),
  "ICA x|x|/2" = list(
    parallel = fastica_update_skew_rankr,
    GS = fastica_update_skew_rankr_gs,
    objective = function(L) colMeans(L * abs(L) / 2)),
  "ICA integrated log-cosh" = list(
    parallel = fastica_update_ilogcosh_rankr,
    GS = fastica_update_ilogcosh_rankr_gs,
    objective = function(L) colMeans(integrated_logcosh(L)))
)
rankr_results <- list()
for (contrast in names(rankr_methods)) {
  method <- rankr_methods[[contrast]]
  for (orth in c("parallel", "GS")) {
    W <- W_rankr
    for (iter in seq_len(n_iter)) W <- method[[orth]](X1, W)
    L <- t(X1) %*% W
    rankr_results[[paste(contrast, orth, sep = " - ")]] <- list(
      loadings = L, objective = method$objective(L))
  }
}
```

```{r rankr-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(rankr_results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "donor")) {
    heading <- if (color_by == "celltype") "Cell type" else "Donor"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(rankr_results[[method]], method, color_by,
                   start_labels = paste0("C", seq_len(r)))
  }
}
```

### rank-r EB-proj with GS

- Fit 25 components from the shared ICA start, using Bernoulli, Exponential,
  and normalized point-Exponential priors.
- Run 50 gradient + 50 Newton warm-up steps with prior $cZ$ and noise variance
  $c$, then 100 Newton steps with prior $Z$ and free precision.
- Apply objective-ordered GS after each direction update, then refresh prior
  parameters and precision. Point-Exponential estimates a separate $q$ for each
  component; its final prior remains centered with unit variance.

```{r rankr-ebproj, cache=TRUE, cache.extra=list(U, W_rankr, tools::md5sum(code_files))}
rankr_eb_results <- list()
for (prior in c("bernoulli", "exponential", "point_exponential")) {
  method <- switch(prior, bernoulli = "Goldplus-Bern - GS",
                   exponential = "Goldplus-Exp - GS", point_exponential = "Goldplus-PointExp - GS")
  rankr_eb_results[[method]] <- eb_proj_goldenplus_rankr_gs(
    U, W_rankr, prior = prior, n_warmup = 50, n_full = 100)
}
```

```{r rankr-ebproj-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(rankr_eb_results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "donor")) {
    heading <- if (color_by == "celltype") "Cell type" else "Donor"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(rankr_eb_results[[method]], method, color_by,
                   start_labels = paste0("C", seq_len(ncol(W_rankr))))
  }
}
```

### SymNMF on the uncentered Gram matrix {.tabset}

I use [Annie's codesymnmf implementation](https://github.com/xxie6/codesymnmf),
copied below, to fit $A = XX^\top/p \approx HH^\top$ with $H\geq0$.
Here `X` is the original simulated matrix, without additional centering or
whitening. I fit 25 components from zeros for 100 iterations. The panels show
all 25 factors.

 <!-- in their original order, each scaled to unit RMS for plotting; -->
<!-- zero factors remain zero. -->

```{r symnmf-code}

codesymnmf_init <- function(A, r, init_type){
  n <- ncol(A)
  if(init_type == 'zeros'){
    H <- matrix(rep(0,n*r), ncol = r)
  }else{
    H <- matrix(runif(n*r), ncol = r)
  }
}

codesymnmf_fit <- function(A, H, maxiter = 100, rand_permutation = FALSE){
  n <- nrow(H)
  r <- ncol(H)

  # scaling (if H is not the zero matrix)
  if (max(H) > 0){
    HHt <- tcrossprod(H)
    beta <- sum(A*HHt)/sum(HHt^2) #scaling factor
    H <- sqrt(beta)*H
  }

  iter <- 1
  x <- rep(0,4)
  # main loop
  while(iter <= maxiter){
    # randomly permute columns if rand_permutation is TRUE
    if (rand_permutation == TRUE){
      col_permutation <- sample(r, size = r, replace = FALSE)
      H <- H[,col_permutation]
    }

    # calculate residual
    R <- A - tcrossprod(H)
    for (k in 1:r){
      R <- R + tcrossprod(H[,k])
      diag_R <- diag(R)
      HtH <- crossprod(H[,k]) #scalar, equals sum(H[,k]^2)
      for (i in 1:n){
        # coefficients a and b of x^3 + ax + b
        HtH <- HtH - H[i,k]^2 #scalar
        a <- HtH - diag_R[i] #scalar
        b <- -(t(H[,k]) %*% R[,i] - H[i,k]*diag_R[i]) #scalar

        Delta <- 4*(a^3) + 27*(b^2) # note: delta can be negative!
        d <- 0.5 * (-b + sqrt(as.complex(Delta/27))) #might be complex

        if (Delta <= 0) {
          r3 <- 2*(abs(d)^(1/3))
          th3 <- atan2(Im(d),Re(d))/3
          x[2] <- r3*cos(th3)
          x[3] <- r3*cos(th3 + (2*pi/3))
          x[4] <- r3*cos(th3 + (4*pi/3))
          x <- x[x >= 0]
          ind <- which.min((x^4)/4 + c(a)*(x^2)/2 + c(b)*x)
          H[i,k] <- x[ind]
          HtH <- HtH + H[i,k]^2
        }else{
          d <- Re(d) # d is just a real number in this case
          z <- sign(d)*(abs(d))^(1/3)
          if (z == 0){ # check if z is zero
            val <- sign(-b)*(abs(b)^(1/3))
          } else {
          val <- z - (a/(3*z))
          }
          if(((val^4)/4 + a*(val^2/2) + b*val < 0) & (val >= 0)){
            HtH <- HtH + val^2
            H[i,k] <- val
          }else{
            H[i,k] <- 0
          }
        }
      } # end of 1:n for loop
      R <- R - tcrossprod(H[,k])
    } # end of 1:r for loop
    iter <- iter + 1
  } # end of maxiter while loop
  # return output
  return(list(H=H, obj_func = sum((A - tcrossprod(H))^2)))
}

codesymnmf <- function(A, r, init_type = 'zeros', maxiter = 100, rand_permutation = FALSE){
  H_init <- codesymnmf_init(A, r, init_type)
  codesymnmf_results <- codesymnmf_fit(A, H_init, maxiter, rand_permutation)
  return(codesymnmf_results)
}
```

```{r symnmf-fit, cache=TRUE, cache.extra=list(X, body(codesymnmf_init), body(codesymnmf_fit), body(codesymnmf))}
A <- tcrossprod(X) / ncol(X)
symnmf_fit <- codesymnmf(A, r = 25, init_type = "zeros", maxiter = 100)
```

```{r symnmf-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
H <- symnmf_fit$H
H_plot <- sweep(H, 2, pmax(sqrt(colMeans(H^2)), 1e-15), "/")
for (color_by in c("celltype", "donor")) {
  heading <- if (color_by == "celltype") "Cell type" else "Donor"
  cat("\n\n#### ", heading, "\n\n", sep = "")
  plot_solutions(list(loadings = H_plot), "SymNMF", color_by,
                 start_labels = paste0("C", seq_len(ncol(H))))
}
```
