---
title: "Simulations: cell types with donor effects"
author: "Junming Guan"
date: "2026-08-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Simulation

Here I aim to look at a simulation with similar strcuture to the pancreas dataset. I simulate the expressions for eight cell types and four donors, with gaussian noise:

$$
X = L_{\mathrm{type}} F_{\mathrm{type}}^\top
  + L_{\mathrm{donor}} F_{\mathrm{donor}}^\top + E.
$$

Every cell type contains all four donors in equal proportions. Donor effects
are shared across cell types. Both loading matrices contain centered indicators.

```{r setup}
set.seed(1)
p <- 1000
donor_sd <- 0.25
noise_sd <- 1

# Counts per donor; the last group stands in for rare cell types.
type_counts <- c(alpha = 222, beta = 120, acinar = 72, ductal = 66,
                 delta = 54, gamma = 30, stellate = 24, rare = 12)
sample_info <- data.frame(
  celltype = factor(rep(rep(names(type_counts), type_counts), 4),
                    levels = names(type_counts)),
  donor = factor(rep(paste0("D", 1:4), each = sum(type_counts)))
)
n <- nrow(sample_info)
table(sample_info$celltype, sample_info$donor)
```

I scale the data to give cell-type variance 1 and donor variance 0.25.

```{r generate-data}
L_type <- scale(model.matrix(~ celltype - 1, sample_info), scale = FALSE)
L_donor <- scale(model.matrix(~ donor - 1, sample_info), scale = FALSE)

F_type <- matrix(rnorm(p * ncol(L_type)), nrow = p)
F_donor <- matrix(rnorm(p * ncol(L_donor)), nrow = p)
F_type <- F_type / sqrt(mean((L_type %*% t(F_type))^2))
F_donor <- donor_sd * F_donor / sqrt(mean((L_donor %*% t(F_donor))^2))

X_type <- L_type %*% t(F_type)
X_donor <- L_donor %*% t(F_donor)
E <- matrix(rnorm(n * p, sd = noise_sd), nrow = n)
X <- X_type + X_donor + E
```

## Check the signal strengths

Donor signal is 6.25% of cell-type signal.

```{r signal-strengths}
data.frame(
  component = c("Cell type", "Donor", "Noise"),
  rms = c(sqrt(mean(X_type^2)), sqrt(mean(X_donor^2)), sqrt(mean(E^2)))
)
sum(X_donor^2) / sum(X_type^2)
```

## PCA and whitening

I retain 25 PCs whereas the the signal has at most 10 dimensions.
<!-- `U` has orthonormal columns; `X1` has PCs in rows and cells -->
<!-- in columns. The signal has at most ten dimensions. -->

```{r pca}
X_centered <- scale(X, scale = FALSE)
n_comp <- 25
svd_fit <- svd(X_centered, nu = n_comp, nv = 0)
U <- svd_fit$u
X1 <- sqrt(n) * t(U)

plot(seq_len(40), svd_fit$d[1:40], type = "b", pch = 16,
     xlab = "PC", ylab = "Singular value", cex.axis = 1.3, cex.lab = 1.3)
abline(v = 10, lty = 2, col = "grey50")
```

## Methods compared

I compare the following rank-1 methods:
- Log-Phi with $\alpha=1$ and $2$
- FastICA with an intercept
- First-order EB-proj with normalized Bernoulli, and centered Exponential.
- Goldenplus EB-proj with normalized Bernoulli, and centered Exponential, bith use ICA-type hessian approximation.

I look at 100 random starts and visualize their solutions.

The normalized Bernoulli is given by

$$
\mathbb{P}[z = \sqrt{(1-p)/p}] = p \quad \mathbb{P}[z = -\sqrt{p/(1-p)}] = 1-p,
$$

where centered Exponential is defined as

$$
Z = T-1,\quad T\sim \text{Exp}(1).
$$

Both have mean $0$ and variance $1$. Note that EB-proj with centered Exp is not really an EB method, as the prior is fixed.

<!-- Use 100 shared random starts, 50 Log-Phi/FastICA updates, and at most 100
updates per EB stage. FastICA uses an intercept; Log-Phi and EB-proj use
centered projections. Bernoulli estimates $\pi$ in
$Z=(B-\pi)/\sqrt{\pi(1-\pi)}$. Exponential fixes $Z=T-1$, $T\sim\mathrm{Exp}(1)$.

For Goldplus-Bern, warm up with prior $cZ$ and fixed noise variance $c$, where
$c=2/(1+\sqrt{5})$, then run Newton with the unit-variance prior $Z$ and free
precision. Both stages use `hessian = "ica"`.
The first-order fits use our `eb_proj_rank1_update()`, initialized with
posterior moments at $\tau=1$ along the same starting directions. -->

```{r method-setup, message=FALSE}
library(ggplot2)
code_files <- file.path("code", c("projection_methods_temp.R",
                                 "ebnm_normalized_bernoulli.R",
                                 "ebnm_centered_exponential.R",
                                 "ebproj_functions.R",
                                 "ebproj_newton_functions.R"))
for (file in code_files) source(file)

set.seed(1)
n_starts <- 100
maxiter <- 100
W_start <- matrix(rnorm(n_comp * n_starts), nrow = n_comp)
W_aug <- rbind(rnorm(n_starts), W_start)
W_start <- sweep(W_start, 2, sqrt(colSums(W_start^2)), "/")
W_aug <- sweep(W_aug, 2, sqrt(colSums(W_aug^2)), "/")
priors <- list(Bernoulli = ebnm_normalized_bernoulli,
               Exponential = ebnm_centered_exponential)
results <- list()
```

```{r ica}
for (alpha in c(1, 2)) {
  W <- W_start
  for (iter in 1:50) W <- fastica_update_logphi(X1, W, alpha = alpha)
  L <- t(X1) %*% W
  results[[paste("Log-Phi alpha =", alpha)]] <- list(
    loadings = L, objective = colMeans(pnorm(alpha * L, log.p = TRUE)))
}

X1_aug <- rbind(1, X1)
W <- W_aug
for (iter in 1:50) W <- fastica_update(X1_aug, W)
L <- t(X1_aug) %*% W
results[["FastICA"]] <- list(loadings = L, objective = colMeans(log(cosh(L))))
```

```{r eb-proj, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
c_golden <- 2 / (1 + sqrt(5))

# Scale posterior moments and density; fitted_g stores the unit prior's shape.
warm_prior <- function(x, s, g_init = NULL, fix_g = FALSE, output = NULL) {
  fit <- ebnm_normalized_bernoulli(x / c_golden, s / c_golden,
    g_init = g_init, fix_g = fix_g, output = output)
  fit$posterior$mean <- c_golden * fit$posterior$mean
  fit$posterior$sd <- c_golden * fit$posterior$sd
  fit$posterior$second_moment <- c_golden^2 * fit$posterior$second_moment
  fit$log_likelihood <- as.numeric(fit$log_likelihood) - length(x) * log(c_golden)
  fit
}

fit_goldplus <- function(w) {
  warm <- eb_proj_newton(U, ebnm_fn = warm_prior, w_init = w, g_init = 0.5,
                        sigma2 = c_golden, fix_tau = TRUE, maxiter = maxiter)
  eb_proj_newton(U, ebnm_fn = ebnm_normalized_bernoulli, w_init = warm$w,
                g_init = warm$fitted_g, tau = warm$tau, maxiter = maxiter)
}

L <- matrix(NA_real_, n, n_starts)
objective <- numeric(n_starts)
status <- character(n_starts)
for (j in seq_len(n_starts)) {
  fit <- fit_goldplus(W_start[, j])
  L[, j] <- fit$x
  objective[j] <- fit$objective
  status[j] <- fit$status
}
eb_results <- list("Goldplus-Bern" = list(loadings = L, objective = objective, status = status))
```

```{r first-order, cache=TRUE, cache.extra=list(U, W_start, n_starts, maxiter, tools::md5sum(code_files))}
P <- tcrossprod(U)
first_order_results <- list()
for (prior_name in names(priors)) {
  prior <- priors[[prior_name]]
  exponential <- prior_name == "Exponential"
  method <- paste0("EBproj-1st-order-", if (exponential) "Exp" else "Bern")
  L <- matrix(0, n, n_starts)
  objective <- numeric(n_starts)
  status <- character(n_starts)
  for (j in seq_len(n_starts)) {
    x_init <- as.vector(t(X1) %*% W_start[, j])
    s_init <- sqrt(n / (n - 2 * n_comp))
    initial <- if (exponential) prior(x_init, s_init) else
      prior(x_init, s_init, g_init = 0.5, fix_g = TRUE)
    fit <- eb_proj_rank1_update(P, initial$posterior$mean, ebnm_fn = prior,
      v2_init = sum(initial$posterior$second_moment),
      k = n_comp, b = n_comp / 2, maxiter = maxiter)
    direction <- as.vector(crossprod(U, fit$vbar))
    direction_norm <- sqrt(sum(direction^2))
    status[j] <- if (fit$niter < maxiter) "converged" else "maximum iterations reached"
    if (fit$v2 == 0) {
      status[j] <- "null solution"
    } else if (direction_norm <= .Machine$double.eps * max(1, sqrt(fit$v2))) {
      status[j] <- "zero projected posterior mean"
    } else {
      L[, j] <- sqrt(n) * U %*% (direction / direction_norm)
    }
    objective[j] <- fit$objective
  }
  first_order_results[[method]] <- list(loadings = L, objective = objective, status = status)
}
```

```{r collect-results}
method_order <- c("Log-Phi alpha = 1", "Log-Phi alpha = 2",
                  "Goldplus-Bern",
                  "EBproj-1st-order-Bern", "EBproj-1st-order-Exp", "FastICA")
results <- c(results, eb_results, first_order_results)[method_order]
```

## Grouped loading plots

Here I group solutions using the method in [Matthew's pancreas note](https://stephenslab.github.io/single-cell-jamboree/pancreas_celseq2_ica.html). For each cluster, the first fit is shown, and the plots are ordered by decreasing objective.
`n` counts starts; `obj` is the representative fit's objective.
Each figure holds up to 25 subplots in five columns, as in Matthew's note,
with fixed subplot sizes and figure height adjusted to the occupied rows.

<!-- EB fits that stop at a null or zero projected -->
<!-- posterior mean are excluded; iteration-limit fits remain in the plots. -->

Cells are ordered by cell type, then donor. Use the tabs under each method
to switch between cell-type and donor coloring.

First-order EB-proj panels plot $L=\sqrt{n}\,Uw$, where
$w=U^\top m/\|U^\top m\|$ and $m$ is the posterior mean before projection.

```{r plot-helper}
plot_solutions <- function(result, method, color_by, start_labels = NULL) {
  L <- result$loadings
  keep <- which(apply(L, 2, sd) > 1e-8 & is.finite(result$objective))
  if (!is.null(result$status)) {
    keep <- keep[result$status[keep] %in% c("converged", "maximum iterations reached")]
  }
  if (!length(keep)) return(invisible(NULL))
  clusters <- if (!is.null(start_labels)) seq_along(keep) else if (length(keep) == 1) 1 else cutree(
    hclust(as.dist(1 - pmin(abs(cor(L[, keep, drop = FALSE])), 1)),
           method = "complete"), h = 0.1)
  groups <- split(keep, clusters)
  representatives <- vapply(groups, function(idx) idx[1], 1L)
  o <- order(result$objective[representatives], decreasing = TRUE)
  representatives <- representatives[o]
  labels <- sprintf("n: %d  obj: %.3f", lengths(groups)[o],
                    result$objective[representatives])
  if (!is.null(start_labels)) labels <- sprintf("%s  obj: %.3f",
    start_labels[representatives], result$objective[representatives])
  cell_order <- order(sample_info$celltype, sample_info$donor)
  pages <- split(seq_along(representatives), ceiling(seq_along(representatives) / 25))
  for (page in seq_along(pages)) {
    idx <- pages[[page]]
    df <- data.frame(
      cell = rep(seq_len(n), length(idx)),
      loading = as.vector(L[cell_order, representatives[idx], drop = FALSE]),
      panel = factor(rep(representatives[idx], each = n), levels = representatives[idx]),
      celltype = rep(sample_info$celltype[cell_order], length(idx)),
      donor = rep(sample_info$donor[cell_order], length(idx)))
    p <- ggplot(df, aes(cell, loading, color = .data[[color_by]])) +
      geom_point(size = 0.5, alpha = 0.7) +
      geom_hline(yintercept = 0, linetype = 2, linewidth = 0.3) +
      facet_wrap(~ panel, ncol = 5, scales = "free_y",
                 labeller = labeller(panel = setNames(labels, representatives))) +
      scale_color_manual(values = hcl.colors(nlevels(df[[color_by]]), "Dark 3")) +
      labs(x = NULL, y = "Loading", color = color_by,
           subtitle = if (length(pages) > 1) paste("Page", page, "of", length(pages))) +
      theme_bw(base_size = 12) +
      theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
            legend.position = "bottom", legend.justification = "left",
            strip.text = element_text(size = 10, margin = margin(2, 0, 2, 0)),
            strip.background = element_rect(fill = "grey90", color = NA)) +
      guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
    g <- ggplotGrob(p)
    panels <- grepl("^panel", g$layout$name)
    g$widths[unique(g$layout$l[panels])] <- grid::unit(1.9, "in")
    g$heights[unique(g$layout$t[panels])] <- grid::unit(1.15, "in")
    g$vp <- grid::viewport(x = 0, y = 1, just = c("left", "top"),
                          width = sum(g$widths), height = sum(g$heights))
    height <- grid::convertHeight(sum(g$heights), "in", valueOnly = TRUE)
    file <- knitr::fig_path(paste0(color_by, "-", make.names(method), "-", page, ".png"))
    xfun::in_dir(knitr::opts_knit$get("output.dir"), {
      dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
      ggsave(file, g, width = 12, height = height + 0.1, units = "in", dpi = 300, bg = "white")
      cat(knitr::knit_hooks$get("plot")(file, knitr::opts_current$get()), "\n\n", sep = "")
    })
  }
}
```

```{r grouped-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "donor")) {
    heading <- if (color_by == "celltype") "Cell type" else "Donor"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(results[[method]], method, color_by)
  }
}
```

## Initialize from the true donor loadings

Here I try initialize EB-proj-NormalizedBernoulli (both first order and goldenplus Newton) from the four donor loadings.

<!-- Project each of the four centered donor indicators in `L_donor` into the
retained PC subspace and normalize its direction. Fit both Bernoulli methods
from these four starts, using the same settings as above. Keep one panel per
starting donor, ordered by decreasing objective. -->

```{r donor-starts, cache=TRUE, cache.extra=list(U, L_donor, maxiter, tools::md5sum(code_files)), dependson="eb-proj"}
W_donor <- crossprod(U, L_donor)
W_donor <- sweep(W_donor, 2, sqrt(colSums(W_donor^2)), "/")
colnames(W_donor) <- levels(sample_info$donor)
donor_results <- list()
for (method in c("EBproj-1st-order-Bern", "Goldplus-Bern")) {
  L <- matrix(0, n, ncol(W_donor))
  objective <- numeric(ncol(W_donor))
  status <- character(ncol(W_donor))
  for (j in seq_len(ncol(W_donor))) {
    if (method == "Goldplus-Bern") {
      fit <- fit_goldplus(W_donor[, j])
      L[, j] <- fit$x
      status[j] <- fit$status
    } else {
      x_init <- as.vector(t(X1) %*% W_donor[, j])
      initial <- ebnm_normalized_bernoulli(x_init, sqrt(n / (n - 2 * n_comp)),
                                         g_init = 0.5, fix_g = TRUE)
      fit <- eb_proj_rank1_update(P, initial$posterior$mean,
        ebnm_fn = ebnm_normalized_bernoulli,
        v2_init = sum(initial$posterior$second_moment),
        k = n_comp, b = n_comp / 2, maxiter = maxiter)
      direction <- as.vector(crossprod(U, fit$vbar))
      direction_norm <- sqrt(sum(direction^2))
      status[j] <- if (fit$niter < maxiter) "converged" else "maximum iterations reached"
      if (fit$v2 == 0) {
        status[j] <- "null solution"
      } else if (direction_norm <= .Machine$double.eps * max(1, sqrt(fit$v2))) {
        status[j] <- "zero projected posterior mean"
      } else {
        L[, j] <- sqrt(n) * U %*% (direction / direction_norm)
      }
    }
    objective[j] <- fit$objective
  }
  donor_results[[method]] <- list(loadings = L, objective = objective, status = status)
}
```

```{r donor-solutions, results='asis', echo=FALSE, fig.width=12, out.width='100%'}
for (method in names(donor_results)) {
  cat("\n\n### ", method, " {.tabset}\n\n", sep = "")
  for (color_by in c("celltype", "donor")) {
    heading <- if (color_by == "celltype") "Cell type" else "Donor"
    cat("\n\n#### ", heading, "\n\n", sep = "")
    plot_solutions(donor_results[[method]], method, color_by,
                   start_labels = colnames(W_donor))
  }
}
```
